Conditional Chemoconnectomics: A Set of Libraries Targeting All Chemical Transmission Corresponding Genes Efficiently

Version 1

Abstract

Abstract

Dissection of neural circuitry underlying behaviors is a central theme in neurobiology. Chemical transmission is the predominant model for signaling between neurons. Here we have created lines target all chemical transmission corresponding genes in Drosophila after modifying GFP RNA interference, Flp-out and CRISPR/Cas9 technologies. After thorough validation, all three strategies are demonstrated to be highly effective with the best using chromatin-peptide fused Cas9 variants and scaffold optimized sgRNAs. As a proof of principle, we conduct a comprehensive intersection analysis of chemoconnectome (CCT) genes expression profiles in the clock neurons using chemoconnectomics driver lines, leading to the finding of 45 CCT genes presented in clock neurons. Mutating these genes specifically in clock neurons revealed that loss of the neuropeptide CNMa in two posterior dorsal clock neurons (DN1p) or its receptor (CNMaR) caused advanced morning activity, opposite to the mutants of neuropeptide PDF or its receptor (PDFR). These results demonstrate the effectiveness of conditional chemoconnectomics libraries and indicate an antagonistic relationship between CNMa-CNMaR and PDF-PDFR signaling in regulating morning anticipation.

Introduction

Much research efforts have been made in uncovering the wiring and signaling pathways of neural circuits underlying specific behaviors. Multiple circuit dissection strategies have been developed, including genetic screening, genetic labelling, circuit tracing, live imaging, genetic sensors and central nervous system reconstruction via electron microscopy (EM) imaging. Recently, we have developed chemoconnectomics (CCTomics), focusing on building a comprehensive set of knockout and knockin tool lines of chemoconnectome (CCT) genes, to dissect neural circuitry based on chemical transmission (Deng et al., 2019).

Each strategy has advantages and disadvantages, for example: genetic screening is less biased but inefficient; circuit tracing with viruses provides information of connection, but is often prone to leaky expression and inaccurate labelling; and EM reconstruction is anatomically elaborate but does not allow for manipulation of corresponding neurons. CCTomics overcomes some limitations of previous strategies by allowing for behavioral screening of CCT genes and accurate labelling or manipulation of corresponding neurons. However, it is still limited in that knockout of some CCT genes can be lethal during development and that CCT genes may function differently in different neurons, which require a cell type-specific manipulation. Thus, we decided to invent a conditional CCTomics (cCCTomics) in which gene deletion was conditional.

There are three major strategies for somatic gene mutagenesis at the DNA/RNA level: RNA interference, DNA site-specific recombination enzymes, and CRISPR/Cas system. RNA interference targets RNAs conveniently and efficiently (Martin and Caplen, 2007; Oberdoerffer et al., 2005). Libraries of transgenic RNAi flies covering almost the entire fly genome have been established (Ni et al., 2011; Perkins et al., 2015). DNA site specific recombination enzymes such as Flp, B3 and Cre mediate specific and efficient gene editing (Gaj et al., 2014; Grindley et al., 2006). These strategies require flies with reverse repetitive sequences knocked into the corresponding genes, which is time-consuming with relatively complex recombination for genetic assays. CRISPR/Cas systems, particularly CRISPR/Cas9, which targets DNA with a sgRNA/Cas protein complex, have been broadly applied in gene manipulation over the last decade. The widespread use of CRIPSR/Cas9 in Drosophila somatic gene manipulation began in 2014 (Xue et al., 2014). Later, tRNA-flanking sgRNAs was designed and applied, which enabled multiple sgRNAs to mature in a single transcript (Xie et al., 2015), accelerating the application of this strategy in conditional gene manipulation in flies with impressive efficiency (Delventhal et al., 2019; Port and Bullock, 2016; Port et al., 2020; Schlichting et al., 2019; Schlichting et al., 2022). Additionally, libraries of UAS-sgRNA targeting kinases (Port et al., 2020) and GPCRs (Schlichting et al., 2022) have been established, but no sgRNA libraries covering all the CCT genes exists yet. The efficiency of CRISPR/Cas9 has not been validated systematically in the Drosophila nervous system.

The circadian rhythm can be used for proof of principle testing of cCCTomics. Organisms evolve periodic behaviors and physiological traits in response to cyclical environmental changes. The rhythmic locomotor behavior of Drosophila, for instance, shows enhanced activity before the light is turned on and off in a light-dark (LD) cycle, referred to as morning and evening anticipations, respectively (Collins et al., 2005; Helfrich-Förster, 2001). Under dark-dark (DD) conditions, the activities peaks regularly about every 24 hours (h) (Konopka and Benzer, 1971). Approximately 150 clock neurons, circadian output neurons and extra-clock electrical oscillators (xCEOs) coordinate Drosophila circadian behaviors (Dubowy and Sehgal, 2017; Tang et al., 2022). The regulation of morning and evening anticipations, the most prominent features in the LD condition, is primarily mediated by four pairs of sLNvs expressing pigment dispersing factor (PDF), six pairs of LNds and the 5th s-LNv (Grima et al., 2004; Rieger et al., 2006; Stoleru et al., 2004). At the molecular level, Pdf and Pdf receptor (PDFR) are well-known, with their mutants showing an advanced evening activity peak and no morning anticipation (Hyun et al., 2005; Lear et al., 2005; Renn et al., 1999). Other neuropeptides and their receptors, including AstC/AstC-R2 and neuropeptide F (NPF) and its receptor (NPFR), have also been reported to modulate evening activities (Díaz et al., 2019; He et al., 2013; Hermann et al., 2012), while CCHa1/CCHa1-R and Dh31 regulate morning activities (Fujiwara et al., 2018; Goda et al., 2019). To date, no advanced morning activity phenotype has been reported in flies.

To develop a more efficient approach for somatic gene manipulation, we have now generated two systems for conditional manipulation of CCT genes. One is GFPi/Flp-out-based conditional knockout system of CCT genes (cCCTomics) and another is CRISPR/Cas9-based (C-cCCTomics). Both systems have achieved high efficiency of gene mutagenesis in the Drosophila nervous system. C-cCCTomics, utilizing chromatin-peptide fused Cas9 and scaffold optimized sgRNA, makes efficient conditional gene knockout as simple as RNAi. Further application of C-cCCTomics in clock neurons revealed novel roles of CCT genes in circadian behavior: CNMa-CNMaR modulates morning anticipation as an antagonistic signal of PDF-PDFR.

Results

Complete Disruption of Target Genes by GFPi and Flp-out Based cCCTomics

For the purpose of conditional chemoconnectomics (cCCTomics), we initially leveraged the benefits of our previously generated CCTomics attP lines (Deng et al., 2019), which enabled us to fuse enhanced GFP coding sequence at the 3’ end of each gene’s coding region and flank most or entire gene span with FRT sequence through site-specific recombination (Fig 1A, Table S1). We designed this system so that it could be used to target genes tagged with GFP by RNAi (Neumüller et al., 2012) and at the same time to enable flippase(Flp) mediated DNA fragment excision between two FRT sequences when FRT sequences are in the same orientation (Vetter et al., 1983)(Fig1A).

cCCTomics Mediates Efficient Conditional Disruption of CCT Genes

(A) Schematic of cCCT gene span and principle of cCCTomics. A T2A-EGFP sequence was induced at the 3’ end of CCT genes and their most or all coding regions (depending on attP-KO lines) were flanked by 34 bp FRT sequence. Both Flp-out (top) and GFP RNAi (down) could mediate CCT gene manipulation. (B-J) Expression of Trh (B-D), CCHa2 (E-G), and Dh31 (H-J) are efficiently disrupted by pan-neuronal expression of GFP-RNAi (B-D), pan-neuronal expression of Flp-out (E-G), and heatshock-Flp (H-J) respectively. Representative fluorescence images of R57C10-Gal4/+;TrhEGFP.FRT/+ (B), UAS-shRNAGFP/TrhEGFP.FRT (C), R57C10-Gal4/+; UAS-shRNAGFP/TrhEGFP.FRT (D), R57C10-Gal4/+;CCHa2EGFP.FRT/+; (E), UAS-Flp/CCHa2EGFP.FRT (F), R57C10-Gal4/+; UAS-Flp/CCHa2EGFP.FRT (G), Dh31EGFP.FRT with heatshock (H), hs-Flp/Dh31EGFP.FRT without heatshock (I), and hs-Flp/Dh31EGFP.FRT with heatshock are shown. Manipulation efficiency and experiment group fly number is noted on the right. Scale bar, 50um. (K-L) sleep profiles (K) and statistical analysis (L) of Flp-out induced nAChRβ2 neuronal knockout flies (red) and genotype controls (dark and blue). Sleep profiles are plotted in 30 min bins. In this and other figures, blank background indicates the light phase (ZT 0-12); shaded background indicates the dark phase (ZT 12-24). Both daytime sleep (open bars) and nighttime sleep (filled bars) duration were significantly reduced in nAChRβ2 neuronal knockout files.

In all statistical panels, unless otherwise noted, 1) Numbers below each bar represent the number of flies tested. 2) Mean ± SEM is shown. 3) The Kruskal-Wallis test followed by Dunn’s post test was used. ***p<0.001. ** p<0.01. * p<0.05. n.s. p>0.05. Male flies were used unless otherwise noted.

To validate the efficiency of cCCTomics, we performed pan-neuronal expression of either shRNAGFP or flipase in cCCT flies. Immunofluorescent imaging showed that constitutive expression of shRNAGFP (Fig 1B-1D, Fig S1A-S1I) or flipase (Fig 1E-1G, Fig S1J-S1U) almost completely eliminated GFP signals of target genes, indicating high efficiency. Knockout at the adult stage using either hsFLP driven Flp-out (Golic and Lindquist, 1989) (Fig 1H-1J) or neurally (elav-Switch) driven shRNAGFP (Nicholson et al., 2008; Osterwalder et al., 2001) (Fig S2A-S2I) also showed high efficiency of suppression. Notably, control group of CCTEGFP.FRT; elav-Switch/UAS-shRNAGFP flies fed with solvent (ethanol) showed obvious decreased GFP (Fig S2B, S2E, and S2H) comparing with UAS-shRNAGFP/CCTEGFP.FRT flies fed with RU486 (Fig S2A, S2D, and S2G), indicating leaky expression of elav-Switch.

We then applied cCCTomics in pan-neuronal knockout of nAChRβ2 which is required for Drosophila sleep (Dai et al., 2021). Ablation of nAChRβ2 in the nervous system dramatically decreased both day and night sleep of flies, which was the same as shown in the previous paper (Dai et al., 2021) (Fig 1K, 1L). Therefore, cCCTomics is an effective toolkit for manipulation of CCT genes and suitable for functional investigations of genes. Expression of in-frame fused eGFP-labelled CCT genes highly co-localized with signals revealed by immunocytochemistry (Fig S3A-S3I), allowing direct examination of gene expression without amplification, which is different from the GAL4/UAS binary system.

We then checked the viability of cCCT lines and found that cCCT lines including CapaEGFP.FRT, ChATEGFP.FRT and EhEGFP.FRT were viable whereas their CCT mutants were lethal. Gad1EGFP.FRT, GluRIIDEGFP.FRT and CapaREGFP.FRT were still lethal as their CCT mutants were (Table S2). This indicates that some of the cCCT knockin flies may functionally affect corresponding genes, which are not suitable for conditional gene manipulation. Combination of cCCT transgenic flies with UAS-flp, UAS-shRNAGFP or specific drivers is complicated and unfriendly for screen work, despite the almost 100% efficiency of gene suppression. Because of the limitations of this method, we further created a CRISPR/Cas9 based conditional knockout system of chemoconnectomics (C-cCCTomics).

CRISPR/Cas9-Based Conditional Knockout System for CCTomics

To simplify effective manipulation of CCT genes, we designed a vector based on pACU2 (Han et al., 2011) with tRNA flanking sgRNAs (Port and Bullock, 2016; Xie et al., 2015) targeting CCT genes (Fig 2A). We also adopted an optimized sgRNA scaffold “E+F” (E, stem extension; F, A-U flip) (Chen et al., 2013), which facilitates Cas9-sgRNA complex formation and gene knockout efficiency (Dang et al., 2015; Poe et al., 2019; Zhao et al., 2016), to all sgRNAs to improve gene knockout efficiency. To balance efficiency and off-target effect, we selected three sgRNAs for each CCT gene with highest predicted efficiency and no predicted off-target effect based on previously reported models (Chu et al., 2016; Doench et al., 2014; Graf et al., 2019; Gratz et al., 2014; Heigwer et al., 2014; Stemmer et al., 2015; Xu et al., 2015) (see Table S3 and details at Methods).

C-cCCTomics Mediates Efficient Conditional Knockout of CCT Genes

(A) Schematic of C-cCCTomics principle. Cas9 and three sgRNAs are driven by GAL4/UAS system. Three tandem sgRNAs are segregated by fly tRNAGly and matured by RNase Z and RNase P. (B-G) Pan-neuronal knockout of Dh31 (D) and Pdf (G) by C-cCCTomics strategy. Representative fluorescence images presented expression of Dh31 (B-D) or anti-PDF (E-G).

Pan-neuronal expression of Cas9 and sgRNA eliminated most (D) or all (G) fluorescent signal compared to control fly brains (B-C, E-F). Scale bar, 50 μm. (H) Activity profiles of pan-neuronal knockout of Pdf. Plotted in 30 min bins. (I-J) Statistical analysis of morning anticipation index (I) and power (J) for pan-neuronal Pdf knockout flies. Knocking out of Pdf in neurons reduced both morning anticipation index and power significantly. (K-L) Statistical analysis of nAChRα2 (K) and nAChRβ2 (L) pan-neuronal knockout flies’ sleep phenotype. Sleep of these flies were not disrupted.

We generated UAS-sgRNA3x transgenic lines for all 209 defined CCT genes (Abruzzi et al., 2017; Dai et al., 2019; Deng et al., 2019) and UAS-Cas9.HC (Cas9.HC) (Mali et al., 2013). We first verified that C-cCCTomics mediated precise target DNA breaking by ubiquitous expression of Cas9.P2 (Port et al., 2014) and sgRNA by targeting Pdf or Dh31. Sanger sequencing showed that indels were present exactly at the Cas9 cleavage sites (Fig S4A-S4H). To determine the efficiency of C-cCCTomics, we employed pan-neuronal expression of Cas9.HC with sgRNADh31 or sgRNApdf. Targeting by Cas9.HC/sgRNADh31 eliminated most but not all of the GFP signal in Dh31EGFP.FRT (Fig 2B-2D), whereas all anti-PDF signals were eliminated by Cas9.HC/sgRNApdf (Fig 2E-2G). Furthermore, we used the C-cCCT strategy to conditionally knockout genes for pdfr, nAChRβ2 and nAChRα2, which were previously reported as essential for circadian rhythm or sleep (Dai et al., 2021; Renn et al., 1999). Pan-neuronal knockout of Pdfr phenocopied arrhythmicity of its null mutants (Fig 2H-2J) while there was no significant sleep decrease in these conditional knockout (cKO) flies (Fig 2K-2L) when we applied C-cCCTomics to manipulate nAChRβ2 or nAChRa2. Taken together, C-cCCTomics (with Cas9.HC) achieved a relatively high gene knockout efficiency, but it was not effective enough for all genes.

Evaluation of Cas9 with Different Chromatin-Modulating Peptides

Since the establishment of the CRISPR/Cas9 system a decade ago, many groups have attempted to improve its efficiency in gene manipulation. Most attempts have been focused on the two main components of this system, the Cas9 protein (Ding et al., 2019; Ling et al., 2020; Liu et al., 2019; Zhao et al., 2016; Zheng et al., 2020) and the single guide RNA (Chen et al., 2013; Chu et al., 2016; Dang et al., 2015; Doench et al., 2014; Filippova et al., 2019; Graf et al., 2019; Labuhn et al., 2018; Mu et al., 2019; Nahar et al., 2018; Scott et al., 2019; Xu et al., 2015). At the beginning of C-cCCTomics design, we adopted an optimized sgRNA scaffold and selected sgRNAs with predicted high efficiency. We tried to further improve the efficacy by modifying Cas9 protein. We fused a chromatin-modulating peptide (Ding et al., 2019), HMGN1 (High mobility group nucleosome binding domain 1), at the N-terminus of Cas9 and HMGB1 (High mobility group protein B1) at its C-terminus with GGSGP linker, termed Cas9.M9 (Fig 3A, Methods). We also obtained a modified Cas9.M6 with HMGN1 at the N-terminus and an undefined peptide (UDP) at the C-terminus (Fig 3A). We replaced the STARD linker between Cas9 and NLS in Cas9.HC with GGSGP the linker (Zhao et al., 2016), termed Cas9.M0 (Fig 3A). None of these modifications have been validated previously in flies.

Efficiency Evaluation of Variations of Chromatin-Modulating Peptides Modified Cas9.

(A) Schematics of chromatin-modulating peptides modified Cas9. (B-J) Efficiency evaluation of Cas9 variants. Fluorescence imaging of R57C10-Gal4>UAS-sgRNAple (B), R57C10-Gal4>UAS-Cas9 (C-F), and R57C10-Gal4>UAS-Cas9, UAS-sgRNAple (G-J) flies are shown. Brains were stained with anti-TH (green). Scale bar is 50μm. (K) Anterior TH positive neurons numbers of (K-U). (L) Statistical analysis of ple knockout efficiency related to (K). Modified Cas9.M6 and Cas9.M9 showed an improved efficiency comparing to Cas9.HC. Student’s t test was used. (M) Statistical analysis of representational KO efficiency of Cas9 variants as related to Figure S5. Gene symbols on the right indicate tested genes. (N) Statistical analysis of Cas9 expression level. (O-P) Workflow of efficiency validation by next-generation sequencing (O) and Statistical analysis of single-site mutation ratios induced by Cas9 variants (P). Paired t test was used in (M), (N) and (P). (Q-R) Statistical analysis of sleep amount for nAChRα2 (X) or nAChRβ2 (Y) pan-neuronal knockout flies. Knockout of nAChRα2 and nAChRβ2 by modified Cas9.M9 significantly decreased flies’ sleep amount.

To determine whether the modified Cas9 variants were more efficient, we first pan-neuronally expressed each Cas9 variant and sgRNAple, and assessed their efficiency by immunofluorescence imaging. By counting anti-TH positive neurons in the brain (anterior view) after targeting by Cas9/sgRNAple, we found that unmodified Cas9.HC/sgRNAple only achieved 69.58±3.04% (n=5) knockout efficiency (Fig 3G, 3K, 3L), while Cas9.M6/sgRNAple and Cas9.M9/sgRNAple significantly improved efficiency to 87.53±3.06% (n=7) and 97.19±2.15% (n=8), respectively (Fig 3I-3L). Fourteen additional CCT genes were subjected to pan-neuronal knockout, and the mRNA levels of the target genes were evaluated using real-time quantitative PCR with at least one primer overlapping the sgRNA targeting site (Fig S5). Cas9.M6 and Cas9.M9 demonstrated significantly higher gene disruption efficiency compared to the unmodified Cas9.HC, achieving average efficiencies of 87.51%±2.24% and 89.59%±2.39% for Cas9.M6 and Cas9.M9, respectively, in contrast to 70.72%±3.82% for Cas9.HC. (Fig 3M, Fig S5). To rule out the possibility of the observed variations in gene disruption efficiency being attributed to differential Cas9 expression levels, we quantified the Cas9 expression levels and noted that both Cas9.M6 and Cas9.M9 exhibited lower mRNA levels than Cas9.HC under the experiment condition (Fig 3N). Subsequently, genomic DNA of Drosophila head was extracted, and libraries encompassing target sites were constructed for high-throughput sequencing to verify disparities in genetic editing efficiency among these three Cas9 variants (Fig 3O). In almost all nineteen sites tested, the mutation ratio consistently showed a trend towards Cas9.M6 and Cas9.M9 having a higher gene disruption efficiency than Cas9.HC (Fig 3P, Fig S6). The single-site mutation rates varied from 5.81% to 43.47% for Cas9.HC, 22.40% to 53.54% for Cas9.M6, and 19.90% to 63.57% for Cas9.M9 (Fig 3P, Fig S6). It should be noted that genomic DNA extracted from fly heads contained glial cells, which did not express Cas9/sgRNA, leading to a larger denominator and consequently reducing the observed mutation rates. Unmodified Cas9 displayed mutation rates comparable to those previously reported by Schlichting et al., 2022. The findings indicated that both Cas9.M6 and Cas9.M9 displayed elevated efficiency compared to Cas9.HC, with Cas9.M9 exhibiting the highest mutagenesis proficiency. These results suggest that the implementation of modified C-cCCTomics using Cas9.M6 and Cas9.M9 achieved an elevated level of efficiency. While unmodified C-cCCTomics was not efficient enough to knockout nAChRβ2 and nAChRα2 to phenocopy their mutants, we employed Cas9.M9 in conditional knockout of these genes to verify its efficiency. Pan-neuronal knocking out of nAChRβ2 or nAChRa2 by Cas9.M9/sgRNA showed significant sleep decrease which was similar to their mutants (Fig 3Q-3R) (Dai et al., 2021).

Taken together, our results support that we have created a high efficiency toolkit for CCT gene manipulation in the nervous system, as well as more efficient Cas9 variants, Cas9.M6 and Cas9.M9, which can also be applied to genes other than those in the CCT.

CCT Genes Found in Clock Neurons by Genetic Intersection

We analyzed the expression profile of CCT genes in circadian neurons with CCTomics driver lines in all clock neurons expressing Clk856 (Gummadova et al., 2009). With the Flp-out or split-lexA intersection strategy (Fig 4A, 4B), we found 45 out of 148 analyzed CCT genes expressed in circadian neurons (Fig 4C, Fig S7-S8, Table S4, Table S5). In all eight subsets of clock neurons, 24 CCT genes were expressed in DN1s, 20 in DN2s, 21 in DN3s, 29 in LNds, 15 in l-LNvs, 11 in s-LNvs, 5 in 5th s-LNv and 3 in LPNs, with a total of 128 gene-subsets.

Genetic Dissection of Clk856 Labelled Clock Neurons.

(A-B) Schematic of intersection strategies used in Clk856 labelled clock neurons dissection, Flp-out strategy (A) and split-LexA strategy (B). (C) Expression profiles of CCT genes in clock neurons. Gradient color denotes proportion of neurons that were positive for the CCT gene within each subset. This figure is corresponding to Table S4.

Prior to transcriptomic analysis (Ma et al., 2021), 23 CCT genes had been reported in clock neurons (Díaz et al., 2019; Duhart et al., 2020; Erion et al., 2016; Frenkel et al., 2017; Fujiwara et al., 2018; Goda et al., 2016; Hamasaka et al., 2007; He et al., 2013; Hermann-Luibl et al., 2014; Hermann et al., 2012; Hyun et al., 2005; Renn et al., 1999; Selcho et al., 2017). In our intersection dissection, expression profiles of 15 CCT genes were similar but not identical to previously reported 18 CCT genes in clock neurons (Fig 4C, Table S4). An additional 22 out of 30 CCT genes identified in our genetic intersection were also detected in the transcriptome analysis (Ma et al., 2021). Among the 45 CCT genes expressed in eight clock neuron subsets, 128 gene-subsets, ∼36% (46 out 128) were found in the transcriptome analysis (Ma et al., 2021) (Fig 4C, Table S4). This suggests that the expression profile obtained through genetic dissection overlapped with transcriptome results (see DISCUSSION).

Conditional Knockout of CCT Genes in Circadian Neurons

To investigate the function of CCT genes in circadian neurons with our conditional knockout system, we knocked out all 67 (45 genes identified above and additional 22 genes reported previously) CCT genes in Clk856-labeled clock neurons by C-cCCTomics.

In the pilot screen, we monitored fly activity by video recording (Dai et al., 2019) and analyzed rhythmic behavior under LD and DD conditions. We analyzed morning anticipation index (MAI) and evening anticipation index (EAI) under the LD condition (Harrisingh et al., 2007; Im and Taghert, 2010)(Fig 5A), power, period and arrhythmic rate (AR) under the DD condition. Fly activities tended to rise rapidly after ZT22.5 at dawn and ZT10 at dusk. Thus, we added two more parameters to describe the anticipatory activity patterns of LD condition. Morning anticipation pattern index (MAPI) was defined as the difference between Pia[arctan(ZT20.5∼ZT22.5 activity increasing slope)] and Pip[arctan(ZT22.5 ∼ZT24 activity increasing slope)], M(Pia-Pip). Evening anticipation pattern index (EAPI) was defined similar to MAPI (Fig 5A, see Methods). Pia and Pip were positive, while MAPI and EAPI were negative, for wild type (wt) flies as their activities gradually increases at dawn or dusk at increasing rates. Knocking out Pdf or Pdfr in clock neurons phenocopied their mutants with lower MAI, advanced evening activity, low power, high arrhythmic rate and shorter period (Hyun et al., 2005; Lear et al., 2005; Renn et al., 1999) (Table S6, Fig S9A-S9D). The MAI-decreasing phenotype of Dh31 knockout was also reproduced in this pilot screen (Goda et al., 2019) (Table S6). All the above results verified the effectiveness of C-cCCTomics. Unexpectedly, additional experimental replications with full controls using Cas9.M9 revealed that leaky expression of Cas9.M9 and sgRNA might have caused disruption of Dh31, Dh44, Pdf and Pdfr (Fig S9A-S9D) (see DISCUSSION), which was not suitable for neuronal specific mutagenesis of some genes. Therefore, in the following work we primarily focused on Cas9.M6 instead.

CNMa Regulation of Morning Anticipation in Clock Neuron

(A) Schematic of MAI, EAI, MAPI and EAPI definition. (B) Activity plots of male flies with CNMa knockout in clock neurons (red) and controls (blue, black and grey), plotted in 30 min bins. An advancement of morning activity peak was presented in CNMa clock-neuron-specific mutants (brown arrowhead). (C-E) Statistical analyses of MAPI, power, and period of flies in (B). MAPI was significantly increased in clock neurons-specific CNMa deficient flies (C) while power (D) and period (E) were not changed. (F) Schematic of CNMaKO generation. The entire encoding region of CNMa was replaced by an attP-SAstop-3P3-RFP-loxP cassette using CRISPR-Cas9 strategy. (G) Statistical analysis of MAPI of male CNMaKO flies (red) and controls (blue and black). MAPI significantly increased in male CNMaKO flies. (H) Activity plots of female CNMaKO flies (red) and controls (blue and black). (I) Statistical analysis of MAPI of female CNMaKO flies (red) and controls (blue and black). MAPI was significantly increased in female CNMaKO flies.

Analysis of the newly defined parameters MAPI and EAPI showed that control flies (Clk856-GLA4>UAS-Cas9.M9) had negative EAPI but slightly positive MAPI. The positive MAPI of control flies in this screen might be caused by Cas9.M9 toxicity. Only the Pdf and Pdfr clock-neuron knockout flies showed positive EAPIs, indicating an advanced evening activity (Table S6, Fig S9A, S9C). nAChRα1, MsR1, mAChR-B, and CNMa cKO flies had highest MAPI values (Table S6). We further confirmed their phenotypes using Cas9.M6 which revealed that CNMa plays a role in regulating morning anticipatory activity (Fig S10A).

Regulation of Morning Anticipation by CNMa-positive DN1p Neurons

Conditionally knocking out CNMa in clock neurons advanced morning activity (Fig 5B, Fig S10B) and increased MAPI (Fig 5C, Fig S10A, S10C), leaving the power and period intact in male flies (Fig 5D, 5E). The same advanced morning activity phenotype were also observed in female flies (Fig S10D-S10G). To further validate this phenotype, we generated a CNMa knockout (CNMaKO) line by replacing its whole coding region with an attP-splicing adaptor element (Deng et al., 2019) (Fig 5F). Both male and female CNMaKO flies exhibited the same phenotypes as seen in the CNMa cKO (Fig 5G-5I).

Previous studies have found CNMa expression in DN1 neurons (Abruzzi et al., 2017; Jin et al., 2021; Ma et al., 2021). Our intersection showed four DN1p and one DN3 CNMa positive neurons in Clk856 labelled neurons (Fig S8 - 16, Fig 4C, Table S4). Analysis with an endogenous CNMa-KI-GAL4 knockin driver showed that six pairs of CNMa neurons located in the DN1p region and three pairs located in the subesophageal ganglion (SOG) had the brightest GFP signals (Fig 6A). The anatomical features of CNMa neurons were further confirmed using stingerRed and more neurons were found in regions, the anterior ventrolateral protocerebrum (AVLP), and the antennal mechanosensory and motor center (AMMC) (Fig S11A). Dendrites of CNMa neurons were concentrated in DN1p and SOG, with their axons distributed around DN1p region, lateral horn (LH), and prow region (PRW) (Fig 6B, 6C). Using the Trans-tango strategy (Talay et al., 2017), we also found that downstream of CNMa neurons were about fifteen pairs of neurons in the SOG, five pairs of LNd neurons, one pair of DN3 neurons and six pars intercerebralis (PI) neurons (Fig 6D, arrowhead).

Expression, Projection and Trans-projection Feature of CNMa Neurons and Its Functional Subset

(A-C) Expression and projection patterns of CNMa-KI-Gal4 in the brain. Membrane, dendrites, and axon projections are labelled by mCD8::GFP (A), Denmark (B), and syt::eGFP (C) respectively. (D) Downstream neurons labelled through trans-tango driven by CNMa-KI-GAL4. Arrowheads indicate candidate downstream neurons: six neurons in PI, one pair in DN3, five pairs in LNd and about 15 pairs in SOG. (E-H) Intersection of DN1p CNMa neurons with DN1p labelled drivers. GMR51H05-GAL4 (E), GMR91F02-GAL4 (F), Pdfr-KI-GAL4 (G) and GMR79A11-GAL4 (H) were intersected with CNMa-p65AD, UAS-LexADBD, LexAop-myr::GFP. Two type I (E, G, H) neurons projected to anterior region and four type II (F) neurons had fewer projections to anterior region. Scale bar, 50μm. (I) MAPIs were significantly increased in all three DN1p drivers mediated CNMa knockout. (J-K) Activity plots of CNMa knockout in R51H05-GAL4 (J) and R91F02-GAL4 (K) neurons. R51H05-GAL4 mediated CNMa knockout flies showed an advanced morning activity peak (J), while R91F02-Gal4 mediated CNMa knockout flies did not (K). (L-M) Statistical analyses of MAI and power. Pdfr reintroduction in R79A11 and R51H05 neurons could partially rescue the MAI-decreased phenotype of Pdfr knockout flies. (N) Statistical analyses of MAI of Pdfr knocking out in CNMa-KI-GAL4 and R51H05-GAL4 labelled neurons.

Because we had found that knocking out CNMa in Clk856-GAL4 labeled neurons produced advanced morning activity, and that CNMa intersected with Clk856-Gal4 labeled neurons in 4 pairs of DN1ps and one pair of DN3 neurons (Fig S8-16), we focused on these neurons and performed more intersections. Taking advantage of a series of clock neuron subset-labeled drivers (Sekiguchi et al., 2020), we intersected CNMa-p65AD with 4 DN1 labelling drivers: GMR51H05-GAL4, GMR91F02-GAL4, Pdfr-KI-GAL4 and GMR79A11-GAL4 (Fig 6E-6H). We found two arborization patterns: Type I with two neurons whose branches projecting to the anterior region, as in CNMa∩GMR51H05, CNMa∩Pdfr, and CNMa∩GMR79A11 (Fig 6E, 5G, 6H), and type II with four neurons branching on the posterior side with few projections to the anterior region, as in CNMa∩GMR91F02 (Fig 6F).

CNMa knockout in Type I or Type II neurons (GMR51H05-GAL4, GMR91F02-GAL4, and GMR79A11-GAL4) all reproduced the MAPI-increased phenotype of clk856 specific CNMa knockout (Fig 6I). However, Type II neurons-specific CNMa knockout (CNMa ∩ GMR91F02) showed no advanced morning activity peak (Fig 6K), while Type I neurons-specific CNMa knockout did (Fig 6J), indicating that these two type I CNMa neurons are the main functional subset regulating the morning anticipation activity of fruit fly.

Pdf or Pdfr mutants have weak or no morning anticipation, which is in reversely related to the phenotype of CNMa knockout flies. We also identified two Pdfr and CNMa double-positive DN1ps, which have a type I projection pattern (Fig 6G). Reintroduction of Pdfr in Pdfr knockout background revealed that GMR51H05 and GMR79A11Gal4 drivers, which covered the main functional CNMa-positive subset, could partially rescue the morning anticipation and power phenotype of Pdfr knockout flies to a considerably larger extent than the GMR91F02 driver (Fig 6L-6M). Moreover, knocking out Pdfr in type I CNMa neurons decreased morning anticipation of flies (Fig 6N). Because the main subset of functional CNMa is also PDFR-positive, it is possible that CNMa secretion is regulated by PDF/PDFR signal.

Role of Neuronal CNMaR in Morning Anticipation

There is only one CNMa receptor reported in the fly genome (Jung et al., 2014). We generated a CNMaRKO-p65AD line by CRISPR/Cas9 (Fig 7A) and this knockout showed advanced morning activity (Fig 7B, 7D) and increased MAPI (Fig 7C, 7E) in both sexes. CNMaRKI-Gal4/UAS-mCD8::GFP and CNMaRKI-Gal4/UAS-stinger::Red showed expression of CNMaR across the whole brain (Fig 7F, Fig S11B), especially in DN1p, DN3, the PI and the SOG. The dendrite arborization and synaptic projections of CNMaR neurons also covered broad regions (Fig 7G, 7H), at the PI, the SOG, the posterior ventrolateral protocerebrum (PVLP) and the central complex (CC). Further conditional knockout of CNMaR in neurons by C-cCCTomics phenocopied CNMaRKO-p65AD phenotype (Fig 7I-7L). These results indicate that CNMaR is similar to CNMa in regulating morning anticipation.

CNMaR Regulation of Morning Anticipation

(A) Schematic of CNMaRKO-p65AD generation. Most of the first exon in CNMaR was replaced by a T2A-p65AD-loxP-3P3-RFP-loxP cassette using CRISPR-Cas9 strategy and the T2A-p65AD was inserted in the reading frame of the remaining CNMaR codon. 3P3-RFP was removed latterly by Cre mediated recombination. (B-E) Activity plot (B, D) and statistical analysis (C, E) of male (B-C) or female (D-E) CNMaRKO-p65AD flies (red) and genotypical controls (blue and black). MAPI was significantly increased in both male and female CNMaRKO-p65AD flies. In this and other figures, “♀” denotes female flies. (F-H) Expression and projection patterns of CNMaR-KI-Gal4 in the brain. Scale bars, 50μm. (I-L) Activity plots (I, K) and statistical analyses (J, L) of CNMaR pan-neuronal knockout flies. Neuronal knockout of CNMaR increased MAPI (K).

Discussion

Conditional CCTomics Strategies and Toolkit

We have generated conditional gene manipulation systems based on Flp-out/GFPi or CRISPR/Cas9. cCCT based gene deletion after heat-shock or mifepristone (RU486) eliminated most GFP signals, and pan-neuronal constitutive expression of shRNAGFP or flippase disrupted seven out of eight tested genes completely while targeting of SIFaEGFP.FRT achieved 96±3% efficiency. Although the recombination of genetic elements is relatively cumbersome when using cCCTomics, it is worthwhile to apply this method to specific genes given its high level of efficiency. While two UAS-sgRNA libraries have been established, one primarily targeting kinases (Port et al., 2020) and the other targeting GPCRs (Schlichting et al., 2022), both libraries only cover a portion of CCT genes, and are thus insufficient for manipulating all CCT genes. The development of C-cCCTomics, however, makes CCT gene manipulation as simple as RNA interference. Furthermore, the use of modified Cas9.M6 or Cas9.M9 highly enhances the efficiency of gene disruption in the nervous system, allowing for efficient manipulation of all CCT genes in a cell-specific manner.

The toxicity of CRISPR/Cas9 depends on the Cas9 protein (Port et al., 2014). When expressed pan-neuronally in nSyb-GAL4 (R57C10-GAL4, attP40), Cas9.M9 slightly reduced viability, while the expression of other Cas9 variants had no significant effect on viability (Fig S12). Although Cas9.M9 showed leaky-expression-efficiency, this was not a problem with Cas9.M6, which successfully disrupted Dh31, Dh44, Pdf, and Pdfr (Fig S9). A more restricted expression of Cas9.M9 with lower toxicity is necessary for better somatic gene manipulation in the future.

CCT of Clock Neurons

Intersecting Clk856-Gal4 or Clk856-p6AD with CCTomics, we identified 45 CCT genes in Clk856 labelled clock neurons. Clock neurons appear highly heterogeneous both in our intersection dissection and in a previous transcriptomic analysis (Abruzzi et al., 2017; Ma et al., 2021). Comparing these two CCT gene expression profiles in clock neurons, 45 out of 128 gene-subsets are identical. The accuracy of our genetic intersection is limited by two possibilities: 1) KI-LexA may not fully represent the expression pattern of the corresponding gene, and 2) the efficiency of STOP cassette removal in the Flp-out strategy is limited. Moreover, the leakage of LexAop-GFP may result in unreliable labelling in split-lexA strategy.

We have also observed that the expression profile of Pdfr in clock neurons is inconsistent across studies, with different clock subsets being identified(Hyun et al., 2005; Im and Taghert, 2010; Lear et al., 2005; Ma et al., 2021; Mertens et al., 2005; Shafer et al., 2008) . The variability in labelling clock neurons with Pdfr transgenic GAL4s, KI-myc tag, antibodies, and transcriptomic anatomy reflects the limitations of each approach. The expression of Pdfr in LNds and DN1s is considered reliable as they are labeled by all strategies. Our intersection dissection is most closely aligned to Pdfr antibody-labelled neurons (Hyun et al., 2005). Both genetic drivers and transcriptomic analysis contribute to our knowledge of the expression profile of neurons. The physiological significance of each gene in a particular neuron should be further investigated by genetic manipulation.

Regulation of Rhythmic Behavior by CCT genes

Multiple attractive genes have been identified in our functional screen of CCT genes in clock neurons: for example, knocking out of VGlut weakens both morning anticipation and rhythmic strength (Table S7). In further screening of brain regions, we have narrowed down the morning anticipation regulation role of VGlut in R18H11-GAL4 labeled neurons (Table S8). VGlut in these neurons has also been reported to regulate sleep in Drosophila (Guo et al., 2016). Its downstream neurons may be the PI neurons or LNvs (Barber et al., 2021; Guo et al., 2016).

Moreover, the deficiency of the neuropeptide CNMa results in advanced morning activity. We have validated that two Pdfr and CNMa double-positive DN1p neurons mainly regulate this process through intersectional manipulation of CNMa. Knockout and reintroduction of Pdfr in these neurons have verified that Pdfr partially functions in DN1p CNMa neurons, and PDF increases cAMP level in Pdfr positive neurons (Shafer et al., 2008), suggesting the regulation of CNMa signaling by PDF signaling. Furthermore, given that the morning anticipation vanishing phenotype of Pdf or Pdfr mutant is opposite to that of CNMa knockout flies, we consider the two signals to be antagonistic. However, knocking out CNMaR in Clk856 labelled clock neurons showed no significant phenotype (Table S7), whereas the mutant and pan-neuronal knockout flies had similar phenotypes to CNMa knockout flies, suggesting its role in the circadian output neurons. Previous studies have indicated that CNMa integrate thermosensory inputs to promote wakefulness, and CNMaR is thought to function in Dh44 positive PI neurons (Jin et al., 2021), a subset of circadian output neurons. To gain a deeper understanding of the downstream effects of DN1p CNMa postive neurons, further analysis focusing on specific brain regions is necessary.

We have also reproduced phenotypes of Pdf, Pdfr, and Dh31 mutant flies with C-cCCTomics as previous studies. Surprisingly, only five genes are functional among all sixty-seven CCT genes in this prior screen. This may be caused by limitations of the simple behavioral paradigm, single gene manipulation, and single GAL4 driver. For example, switching of light condition from L: D = 12h: 12h to L: D = 6h:18h, AstC/AstC-R2 would suppress flies’ evening activity intensity to adapt to the environmental change (Díaz et al., 2019), and only double knockout of AChRs and mGluRs in PI neurons can possibly result in alteration in behavioral rhythms (Barber et al., 2021). Further diversified functional analysis of CCT genes in clock neurons is required for clock circuit dissection.

Methods

Fly Lines and Rearing Conditions

Flies were reared on standard corn meal at 25 ℃, 60% humidity, 12 h light: 12 h dark (LD) cycle. For flies used in behavior assays, they were backcrossed into our isogenized Canton S background for 5 to 7 generations. For heat induced assays, flies were reared at 20℃. All CCT attP KO lines and CCT KI driver lines were previous generated at our lab (Deng et al., 2019).

Clk856-GAL4 and GMR57C10-GAL4 driver lines were gifts from Donggen Luo Lab (Peking University). 13XLexAop2 (FRT.stop) myr::GFP was gift from Rubin Lab.

C-cCCTomics sgRNAs design

All sgRNAs target at or before functional coding regions (e.g. GPCR transmembrane domain, synthetase substrate binding domain) of each CCT genes. For each gene, about 20 sgRNAs with specific score ≥ 12 were firstly designed at CRISPRgold website (Chu et al., 2016; Graf et al., 2019), then their specificity and efficacy were further valued in Optimal CRISPR target finder(Gratz et al., 2014), E-CRISPR(Heigwer et al., 2014) and CCTop (Stemmer et al., 2015) system. The first three highest efficacy sgRNAs with no predicted off-target effect were selected. All selected sgRNAs are listed in Table S3.

Molecular Biology

All cCCTomics knockin (KI) lines and C-cCCTomics transgenic flies were generated through phiC31 mediated attB/attP recombination, and the miniwhite gene was used as selection marker.

For cCCTomics KI lines, backbone pBSK-attB-FRT-HpaI-T2A-EGFP-FRT was modified from pBSK-attB-loxP-myc-T2A-Gal4-GMR-miniwhite (Deng et al., 2019). Myc-T2A-GAL4 cassette was removed by PCR amplification while first FRT cassette was introduced. Second FRT cassette was inserted by T4 ligation between SpeI and BamHI. T2A-EGFP was cloned from pEC14 and was inserted into the backbone between two FRT cassettes. All gene spans, except for stop codon, deleted in CCT attP KO lines were cloned into pBSK-attB-FRT-HpaI-T2A-EGFP-FRT at HpaI site.

For C-cCCTomics UAS-sgRNA lines, backbone pMsgNull was modified from pACU2 (Han et al., 2011). Synthetic partial fly tRNAGly sequence was inserted between EcoRI and KpnI. An irrelevant 1749 bp cassette amplified from pAAV-Efla-DIO-mScarlet (addgeen#130999) was inserted between EagI and KpnI. All sgRNA spacers were synthesized at primers, and “E+F” sgRNA scaffold and rice tRNAGly was amplified from a synthetic backbone PM04. Finally, gRNA-tRNAGly cassettes were cloned into pMsgNull between EagI and KpnI by Gibson Assembly.

All UAS-Cas9 variants generated in this research were cloned into vector pACU2 and all Cas9 sequence were amplified from hCas9 (addgene#41815). Human codon optimized Cas9 was cloned into pACU2 to generate UAS-Cas9.HC. UAS-Cas9.M0 was modified from UAS-Cas9.HC by introduce a 3x HA tag after NSL and replace the SARD linker with the 3×GGSGP linker (Zhao et al., 2016). UAS-Cas9.M6 and UAS-Cas9.M9 were designed as HMGN1-Cas9-UPD and HMGN1-Cas9-HB1 respectively. All these chromatin-modulating peptides were linked with Cas9 by 3x GGSGP linker.

CNMaRKO-p65AD is generated by replace the coding region of the first exon with T2A-p65AD by CRISPR/Cas9, the T2A-p65AD was linked in frame after first ten amino acids. Spacers of gRNAs used to break the targeted CNMaR region were 5’-GCAGATTTCAGTTCATCTTT-3’, 5’-GGCTTGGCAATGAC TATATA-3’.

Gene expression quantitation and high-throughput sequencing

Female flies were gathered six to eight days post-eclosion for gene expression quantification and high-throughput sequencing. Fly heads were isolated by chilling them on liquid nitrogen and subsequent shaking. mRNA extraction was performed using Trizol according to a previously established protocol (Green and Sambrook, 2020). Genomic DNA was removed, and cDNA was synthesized using a commercial kit (TIANGEN#DP419). For real-time quantitative PCR, at least one PCR primer was designed to overlap with the sgRNA target site.

Genomic DNA from fly heads was extracted using a standard alkali lysis protocol(Huang et al., 2009). Genomic regions approximately 130 to 230 bp in length, centered around the sgRNA target site, were amplified by PCR employing Q5 polymerase (NEB#M0494). Subsequently, libraries were prepared using the BTseq kit (Beijing Tsingke Biotech Co., Ltd.). These libraries were pooled and subjected to sequencing on the MiSeq platform (Illumina). Analysis of the libraries was conducted using Crispresso2(Clement et al., 2019).

Generation of KI and Transgenic lines

Generation of cCCTomics KI, CNMaKI-p65AD, and CNMaRKO-p65AD lines are the same as generation of CCTomics KI driver lines as previously described (Deng et al., 2019). To generate C-cCCTomics UAS-sgRNA or UAS-Cas9 variants lines, attB vectors were injected and integrated into the attP40, attP2 or VK00005 through phiC31 mediated gene integration.

All flies generated in this research were selected by mini-white and confirmed by PCR.

Behavioral assays

Unmated male or female flies of 4 to 5 days were used in circadian rhythm assays. Before measurement, flies were entrained under 12 h light: 12 h dark cycle at 25℃ for at least 3 days and then transferred to dark-dark condition for 7 days.

Virgin flies of 4 to 5 days were used in sleep assays. Flies were entrained to a 12 h light: 12 h dark cycle at 25℃ for 2 days to eliminate the effect of CO2 anesthesia before sleep record. Sleep was defined as 5 min or longer immobility (Hendricks et al., 2000; Shaw et al., 2000) and analyzed by in-house scripts as previously described ((Dai et al., 2021; Dai et al., 2019; Deng et al., 2019).

Locomotion was obtained as previously described (Dai et al., 2021). Locomotion activity was measured and analyzed by Actogram J plugin (Dai et al., 2019). Morning anticipation index and evening anticipation index were defined as the ratio of last 3 hours activity before light-on or light-off accounts to last 6 hours activity before light-on or light-off (Index=sum(3hrs)/sum(6hrs)) (Harrisingh et al., 2007; Im and Taghert, 2010) and analyzed by an in-house python script.

Heat Shock and Drug Treatment

For hsFLP mediated conditional knockout, flies of 4 to 6 days were heat shock at 37℃ during ZT10 to ZT12 for 4 days. And they were reared at 20℃ for another 4-days and then dissected.

For mifepristone (RU486) induced conditional knockout, flies of 4 to 6 days were treated with 500 μM RU486 mixed in corn food and then dissected 4 days later.

Immunohistochemistry and Confocal Imaging

For all imaging without staining, adult flies were anesthetized on ice and dissected in cold phosphate-buffer saline (PBS). Brains or ventral nerve codes (VNC) were fixed in 2% paraformaldehyde (weight/volume) for 30 min, washed with washing buffer (PBS with 1% Triton X-100, v/v, 3% NaCl, g/mL) for 7 min three times and mounted in Focusclear (Cell Explorer Labs, FC-101).

For imaging with staining, brains and VNCs were fixed for 30 min, washed for 15 min three times. Then they were blocked in PBSTS, incubated with primary antibodies, washed with washing buffer, incubated with second antibodies and mounted as described previously(Dai et al., 2021; Dai et al., 2019).

All brains or VNCs were imaged on Zeiss LSM710 or Zeiss LSM880 confocal microscope and processed by Imaris.

The following primary antibodies were used: mouse anti-PDF (1:200, DSHB), rabbit anti-TH (1:1000, Novus Biologicals), rabbit anti-LK (1:1000, RaoLab, this paper). Rabbit anti-DSK (1:1000) was a gift from Dr. C. Zhou Lab (Institute of Zoology, Chinese Academy of Science) (Wu et al., 2020). The following secondary antibodies were used: AlexaFluor goat anti-mouse 488 (1:1000, Invitrogen), AlexaFlour goat anti-rabbit 488/633 (1:1000, Invitrogen).

For Fig 2, number of TH positive neurons were counted with Imaris Spots plugin.

Quantification and Statistics

MAI, MAPI, EAI, EAPI, power and period were calculated by python or R scripts. ZT0 was set as the time point when light was on and ZT12 was set as the time point for light-off. Activity bins started at ZT0 and each was calculated as a sum of the total activity within 30min. Flies were regarded as death and removed if their activity value within last 2 bins was 0. A representative 24hr activity pattern was the average between corresponding activity bins from 2 consecutive days. To minimize effects from singular values, each flies‣ activity was normalized using the following formula: bi means the activity value for a certain bin. min(b0,…,b48) means the minimal bin value within 24hr and max(b0,…,b48) means the maximal bin value within 24hr. Nor_bi means the final normalized bin value for a certain bin from a given fly. Normalized activity was used for following analysis.

Morning activity arise (M_arise) was defined as the radian between the activity curve (ZT21-ZT22) and the time coordinate. Morning activity plateau (M_plateau) was defined as the radian between the activity curve (ZT22.5-ZT24) and the time coordinate. Evening activity arise (E_arise) was defined as the radian between the activty curve (ZT8-ZT11.5) and the time coordinate. Evening activity plateau (E_plateau) was defined as the radian between the activity curve (ZT10.5-ZT12) and the time coordinate. Morning anticipation pattern index (MAPI) was calculated by subtracting M_arise from M_plateau and Evening anticipation pattern index (EAPI) was calculated by subtracting E_arise from E_plateau.

The original activity data from 7 consecutive days in dark-dark condition was used for power and period calculation as described (Geissmann et al., 2019). Each flies‣ periodogram was calculated based on Chi-Square algorithm (Sokolove and Bushell, 1978) and flies with a null power value were regarded as arrhythmic.

All statistical analyses were carried out with Prism 8 (Graphpad software). The Kruskal-Wallis ANOVA test followed by Dunn’s posttest was used to compare multiple columns.

eLife Assessment

The reviewers were generally enthusiastic about this study, noting that it reports new tools that could be valuable to the community. However, the results in follow-up experiments testing specific tools are often inconsistent with those in the literature, and some conclusions are overstated. More details are needed for both sections, the tool generation and the testing of the CNMa peptide. While the evidence for the tool development was solid, evidence demonstrating the validity of the tools was deemed incomplete.

Reviewer #1 (Public Review):

Summary:

The paper of Mao et al. expands the genetic toolset that was previously developed by the Rao lab (Denfg et al 2019) to introduce the conditional KO or downregulation of neurotransmission components in Drosophila. The authors then use these tools to investigate neurotransmission in the clock neurons of the Drosophila brain. They first test some known components and then analyze the contribution of the CNMa neuropeptide and its receptor to the circadian behavior. The results indicate that CNMA acts from a subset of DN1ps (dorsal clock neurons) to set the phase of the morning peak of locomotor activity in light:dark cycles, with an advanced morning activity in the absence of the neuropeptide. Interestingly, the receptor for the PDF neuropeptide appears to be acting in some of the CNMa neurons to control morning activity.

Strengths/weaknesses:

This is clearly a very useful new set of tools to restrict the manipulation of these components to specific neuronal populations, and overall (see specific points below), the paper is convincing to show that the tools indeed allow to efficiently and specifically eliminate neuropeptides/receptors from subsets of neurons. The analysis of the CNMa function in the clock network reveals a new and interesting function for CNMa. but this part needs to be improved. Some of the behavioral data (PDF/PDFR) do not fit with published work with the mutants. This should be clarified by providing more data comparing the described genotypes with the classical mutants. Some conclusions also need to be toned down.

Reviewer #2 (Public Review):

In this study, Mao and co-workers deliver a substantial suite of genetic tools in support of the senior author's recent proposal to create a "chemoconnectomic" tool kit for the expression mapping and conditional disruption of specific neurotransmitter systems with fly neurons of interest. Specifically, they describe the creation of two toolsets for recombination-based and CRISPR/Cas9-based conditional knockouts of genes supporting neurotransmitter and neuromodulator function and Flp-Out and Split-LexA toolkit for the examination of gene expression within defined subsets of neurons. The authors report the creation of conditional genetic tools for the disruption/mapping of approximately 200 chemoconnectomic gene products, an examination of the general effectiveness of these tools in the fly brain, and apply them to the circadian clock network in an attempt to reveal new information regarding the transmitter/modulator systems involved in daily behavioral timing. The authors provide clear evidence of the effectiveness of the new methods along with a transparent assessment of the variability of the tools. In addition, they present evidence that the neuro peptide CNMa influences the morning peak of daily activity in the fly by regulating the timing of activity increases in anticipation of dawn.

A major strength of the study is the transparent assessment of the effectiveness and variability of the conditional genetic approaches developed by the authors. The authors have largely achieved their aims and the study therefore represents a major delivery on the promise of chemoconnectomics made by the senior author in 2019 (Neuron, Vol. 101, p. 876). Though there are some concerns about the variability of knockout effectiveness, off-target effects of the knockout strategies, and (especially) the accuracy of the gene expression approach, the tools created for this study will almost certainly be useful for the field and support a great deal of future work.

Reviewer #3 (Public Review):

Summary:

Mao and colleagues generated powerful reagents to genetically analyse chemical communication (CCT) in the brain, and in the process uncovered a function for the CNMa neuropeptide expressed in a subset of DN1p neurons that contributes to the temporal organization of locomotor activity, i.e., the timing of morning anticipation.

Strengths:

The strength of the manuscript relies on the generation/characterization of new tools for conditional targeting a well-defined set of CCT genes along with the design and testing of improved versions of Cas9 for efficient knockout. Such invaluable resources will be of interest to the whole community. The authors employed these tools and intersectional genetics to provide an alternative profiling of clock neurons, which is complementary to the ones already published. Furthermore, they uncovered a role for CNMamide, expressed in two DN1ps, in the timing of morning anticipation.

Weaknesses:

They targeted an extensive set of candidate genes putatively involved in communication (transporters, receptor subunits, neuropeptides, neurotransmitter synthesis, etc); they provide a list of efficient gRNAs to target even a longer list of candidate genes, however, it is not clear if all of those made it into transgenic lines that effectively mediate targeting all chemical transmission genes (as suggested by the authors).

Author Response

The following is the authors’ response to the original reviews.

We are grateful to the reviewers for their constructive comments. The following is our point-to-point responses.

Reviewer #1 (Recommendations For The Authors):

Point 1- Abstract: advanced morning peak « opposite » to pdf/pdfr mutants. To my knowledge, the alteration of PDF/PDFR suppresses the morning peak. I am not sure that an advance of the peak is « opposite » to its inhibition?

Mutants with disruptions in CNMa or CNMaR display advanced morning activity, indicating an enhanced state. Mutants with disruptions in Pdf or Pdfr exhibit no morning anticipation, suggesting a promoting role of these genes in morning anticipation. Therefore, our revised version is: “Specific elimination of each from clock neurons revealed that loss of the neuropeptide CNMa in two posterior dorsal clock neurons (DN1ps) or its receptor (CNMaR) caused advanced morning activity, indicating a suppressive role of CNMa-CNMaR on morning anticipation, opposite to the promoting role of PDF-PDFR on morning anticipation.” (Line 43-51)

Point 2- Fig 1K-L: the authors should show the sleep phenotype of the homozygous nAChRbeta2 mutant (if not lethal) for a direct comparison with the FRT/FLP genotype and thus evaluate the efficiency of the system.

We have incorporated sleep profiles of nAChRbeta2 mutant and W1118 into Fig 1K-L. nAChRbeta2 mutants (red) exhibited a sleep amount comparable to that of pan-neural nAChRbeta2 knockout flies (dark red), as shown below.

Point 3- Dh31-EGFP-FRT expression patterns look different in figS1 A (or fig1 H) and J. why that?

We re-examined the original data. Both (with R57C10-GAL4 for Fig. S1A, right, S1J, left) are Dh31EGFP.FRT samples displayed below which demonstrated consistent primary expression subsets. Any observed disparities in region "e" could potentially be attributed to variations during dissection.

Point 4- The knockdown experiments with the elav-switch (RU486) system (fig S2) do not seem to be as efficient as the HS-FLP system (fig 1H-J). The conclusions on the efficiency should be toned down.

We have revised accordingly: "Near Complete Disruption of Target Genes by GFPi and Flp-out Based cCCTomics" (Line 130): "Knocking out at the adult stage using either hsFLP driven Flp-out (Golic and Lindquist, 1989) (Fig. 1H-1J) or neural (elav-Switch) driven shRNAGFP (Nicholson et al., 2008; Osterwalder et al., 2001) (Fig. S2A-S2I), also resulted in the elimination of most, though not all, GFP signals." (Line 145-149)

Point 5- Fig 2H-J: the LD behavioral phenotype of pdfr pan-neuronal cripsr does not seem to correspond to what is described in the literature for the pdfr mutant (han), see hyun et al 2005 (no morning anticipation and advanced evening peak). I understand that the activity index is lower than controls but fig2H shows a large anticipatory activity that seems really unusual, and no advanced evening peak is observed. I think that the authors should show the CRISPR flies and pdfr mutants together, to better compare the phenotypes.

Thank you for pointing out that the phenotypes of pan-neuronal knockout of PDFR by unmodified Cas9 (Fig. 2H-2I of the previous version) whose morning anticipation still exist (Fig, 2H of the previous manuscript), although the significant decrease of morning anticipation index (Fig 2I of the previous manuscript) and advanced evening activity are not as pronounced as observed in han5304 (Fig. 3C in Hyun et al., 2005).

First, we have separated the activity plots of Fig. 2H of previous manuscript, as shown below. The activity from ZT18 to ZT24 shows a tendency of decreasing from ZT18 to ZT21 and a tendency of increasing from ZT21 to ZT24. The lowest activity before dawn during ZT18 to ZT24 shows at about ZT21, and the activity at ZT18 is comparable to the activity at ZT24. This is significantly different compared to the two control groups, whose activity tends to increase activity from ZT18 to ZT24 with an activity peak at ZT24.

The activity from ZT6 to ZT12 increased much faster in Pdfr knockout flies and get to an activity plateau at about ZT11 compared to two control groups with a slower activity increasing from ZT6 to ZT12 with no activity plateau but an activity peak at ZT12.

Second, we have incorporated the phenotype of Pdfr mutants we previously generated (Pdfr-attpKO Deng et al., 2019) with Pdfr pan-neuronal knockout by Cas9.HC. This mutant lacks all seven transmembrane regions of Pdfr (a). The phenotypes are very similar between Pdfr-attpKO flies and Pdfr pan-neuronal knockout flies. In this experimental repeat, we found that a much more obvious advanced evening activity peak is observed both in pan-neuronal knockout flies and Pdfr-attpKO flies.

To further analyze the phenotypes of Pdfr pan-neuronal knockout flies by Cas9.HC, we referred to the literature. The activity pattern at ZT18 to ZT24 (activity tends to decrease from ZT18 to ZT21 and tends to increase from ZT21 to ZT24, with the lowest activity before dawn occurring at about ZT21, and activity at ZT18 comparable to activity at ZT24) is also reported in Pdfr knockout flies such as Fig3C and 3H in Hyun et al., 2005, Fig 2B in Lear et al., 2009, Fig 3B in Zhang et al., 2010, Fig .5A in Guo et al., 2014, and Fig 5B in Goda et al., 2019. Additionally, the less pronounced advanced evening activity peak compared to han5304 (Fig. 3C in Hyun et al., 2005) is also reported in Fig. 2B in Lear et al., 2009, Fig. 3B in Zhang et al., 2010, and Fig. 5B in Goda et al., 2019. We consider that this difference is more likely to be caused by environmental conditions or recording strategies (DAM system vs. video tracing).

Therefore, we revised the text to: “Pan-neuronal knockout of Pdfr resulted in a tendency towards advanced evening activity and weaker morning anticipation compared to control flies (Fig. 2H-2I), which is similar to Pdfr-attpKO flies. These phenotypes were not as pronounced as those reported previously, when han5304 mutants exhibited a more obvious advanced evening peak and no morning anticipation (Hyun et al., 2005)”.

Point 6-The authors should provide more information about the DD behavior (power is low, but how about the period of rhythmic flies, which is shortened in pdf (renn et al) and pdfr (hyun et al) mutants).

We have incorporated period data into Fig. 2I. Indeed, conditional knock out of Pdfr by Cas9.HC driven by R57C10-GAL4 shortens the period length, as shown below (previous data), also in Fig. 2I of the revised version.

In the revised Fig. 2I, we tested 45 Pdfr-attpKO flies during DD condition (3 out of 48 flies died during video tracing in DD condition), and only one fly was rhythmic. In contrast, 9 out of 48 Pdfr pan-neuronal knockout flies were rhythmic.

Point 7- P15 and fig6. The authors indicate that type II CNMa neurons do not show advanced morning activity as type I do, but Figs 6 I and K seem to show some advance although less important than type I. I am not sure that this supports the claim that type I is the main subset for the control of morning activity. This should be toned down.

We have re-organized Fig. 6 and revised the summary of these results as: “However, Type II neurons-specific CNMa knockout (CNMa ∩ GMR91F02) showed weaker advanced morning activity without advanced morning peak (Fig. 6N), while Type I neurons-specific CNMa knockout did (Fig. 6J), indicating a possibility that these two type I CNMa neurons constitute the main functional subset regulating the morning anticipation activity of fruit fly”. (Line 400-405)

Point 8- Figs 6M and N: is power determined from DD data? if yes, how about the period and arrhythmicity? Please also provide the LD activity profiles for the mutants and rescued pdfr genotypes.

Yes, the power was determined from the DD data. In the new version of the manuscript, we have included the activity plots for the LD phase in supplementary Fig S13, as well as shown below (A, B), and the period and arrhythmicity data for the DD phase in Fig. 6S and Table S7. We have also refined the related description as follows: “Moreover, knocking out Pdfr by GMR51H05, GMR79A11 and CNMa GAL4, which cover type I CNMa neurons, decreased morning anticipation of flies (Fig. 6T, Fig. S13B). However, the decrease in morning anticipation observed in the Pdfr knockout by CNMa-GAL4 was not as pronounced as with the other two drivers. Because the presumptive main subset of functional CNMa is also PDFR-positive, there is a possibility that CNMa secretion is regulated by PDF/PDFR signal”. (Line 413-419)

Point 9- Fig 7: does CNMaR affect DD behavior? This should be tested.

We analyzed the CNMaR-/- activity in the dark-dark condition over a span of six days. Results revealed a higher power in CNMaR mutants compared to control flies (Power: 93.5±41.9 (CNMaR-/-, n=48) vs 47.3±31.6 (w1118, n=47); Period: 23.7±0.3 h (CNMaR-/-, n=46) vs 23.7±0.3 h (w1118, n=47); arrhythmic rate 2/48 (CNMaR-/-) vs 0/47 (w1118)). Considering that mutating CNMa had no obvious effect on DD behavior, even if CNMaR affects DD behavior, it cannot be attributed to CNMa signal, we did not further repeat and analyze DD behavior of CNMaR mutant. We believe this raises another question beyond the scope of our current discussion.

Reviewer #2 (Recommendations For The Authors):

Point 1-One major concern is the apparent discrepancies in clock network gene expression using the Flp-Out and split-LexA approaches compared to what is known about the expression of several transmitter and peptide-related genes. For example, it is well established that the 5th-sLNv expresses CHAT (along with a single LNd), yet there appears to be no choline acetyltransferase (ChAT) signal in the 5th-sLNv as assayed by the Split-LexA approach (Fig. 4). This approach also suggests that DH31 is expressed in the s-LNvs, which, as one of the most intensely studied clock neuron are known to express PDF and sNPF, but not DH31. The results also suggest that the sLNvs express ChAT, which they do not. Remarkably PDF is not included in the expression analysis, this peptide is well known to be expressed in only two subgroups of clock neurons, and would therefore be an excellent test case for the expression analysis in Fig. 4. PDF should therefore be added to analysis shown in Fig. 4. Another discrepancy is PdfR, which split LexA suggests is expressed in the Large LNvs but not the small LNvs, the opposite of what has been shown using both reporter expression and physiology. The authors do acknowledge that discrepancies exist between their data and previous work on expression within the clock network (lines 237 and 238). However, the extent of these discrepancies is not made clear and calls into question the accuracy of Flp-Out and Split LexA approaches.

The concerns mentioned above are:

(1) sLNvs express PDF and sNPF but not Dh31;

(2) ChAT presents in 5th-sLNv and one LNd but not in other sLNvs;

(3) PDFR presents in sLNvs but not l-LNvs.

(4) PDF is not included in the analysis.

To verify the accuracy of these intersection analyses, all related to PDF positive neurons (except 5th-sLNv and LNds), we stained PDF and examined the co-localization between PDF-positive LNvs and the respective drivers ChAT-KI-LexA, Pdfr-KI -LexA, Dh31-KI -LexA, and Pdf-KI -LexA.

First, Dh31-KI-LexA labeled four s-LNvs, as shown below (also in Fig. S9A). Therefore, the results of the intersection analysis of Dh31-KI-LexA with Clk856-GAL4 are correct. The difference in the results compared to previous literature is attributed to Dh31-KI-LexA labels different neurons than the previous driver or antibody.

Second, no s-LNv was labeled by ChAT-KI -LexA as shown below. We rechecked our intersection data and found that we analyzed 10 brains of ChAT-KI-LexA∩Clk856-GAL4 while only two brains showed sLNvs positively. To enhance the accuracy of intersection analysis results, we marked all positive signal records when positive subsets were found in less than 1/3 of the total analyzed brains (Table S4).

Third, one l-LNv and at least two s-LNvs were labeled by Pdfr-KI-LexA, as shown below (also in Fig. S9B). Fourth, Pdf-KI-LexA labels all PDF-positive neurons, but the intersection analysis by Pdf-KI-LexA and Clk856-GAL4 only showed scattered signals, as shown below (D, also in Fig. S9C). For these cases, we found some positive signals expected but not observed in our dissection. The possible reason could be the inefficiency of LexAop-FRT-myr::GFP driven by LexA. Therefore, our intersection results must miss some positive signals.

Finally, we revised the text to (Line 286-317):

To assess the accuracy of expression profiles using CCT drivers, we compared our dissection results with previous reports. Initially, we confirmed the expression of CCHa1 in two DN1s (Fujiwara et al., 2018), sNFP in four s-LNvs and two LNds(Johard et al., 2009), and Trissin in two LNds (Ma et al., 2021), aligning with previous findings. Additionally, we identified the expression of nAChRα1, nAChRα2, nAChRβ2, GABA-B-R2, CCHa1-R, and Dh31-R in all or subsets of LNvs, consistent with suggestions from studies using ligands or agonists in LNvs (Duhart et al., 2020; Fujiwara et al., 2018; Lelito and Shafer, 2012; Shafer et al., 2008) (Table S4).

Regarding previously reported Nplp1 in two DN1as (Shafer et al., 2006), we found approximately five DN1s positive for Nplp-KI-LexA, indicating a broader expression than previously reported. A similar pattern emerged in our analysis of Dh31-KI-LexA, where four DN1s, four s-LNvs, and two LNds were identified, contrasting with the two DN1s found in immunocytochemical analysis (Goda et al., 2016). Colocalization analysis of Dh31-KI-LexA and anti-PDF revealed labeling of all PDF-positive s-LNvs but not l-LNvs (Fig S9A), suggesting that the differences may arise from the broader labeling of 3' end knock-in LexA drivers or the amplitude effect of the binary expression system. The low protein levels might go undetected in immunocytochemical analysis. This aligns with transcriptome analysis findings showing Nplp1 positive in DN1as, a cluster of CNMa-positive DN1ps, and a cluster of DN3s (Ma et al., 2021), which is more consistent with our dissection.

Despite the well-known expression of PDF in LNvs and PDFR in s-LNvs (Renn et al., 1999; Shafer et al., 2008), we did not observe stable positive signals for both in Flp-out intersection experiments, although both Pdf-KI-LexA and Pdfr-KI-LexA label LNvs as expected (Fig S9B-S9C). We also noted fewer positive neurons in certain clock neuron subsets compared to previous reports, such as NPF in three LNds and some LNvs (Erion et al., 2016; He et al., 2013; Hermann et al., 2012; Johard et al., 2009; Lee et al., 2006) and ChAT in four LNds and the 5th s-LNv (Johard et al., 2009; Duhart et al., 2020) (Table S4). We attribute this limitation to the inefficiency of LexAop-FRT-myr::GFP driven by LexA, acknowledging that our intersection results may miss some positive signals.

Point 2-Related to this, the authors rather inaccurately suggest that the field's understanding of PdfR expression within the clock neuron network is "inconsistent" and "variable" (lines 368-377). This is not accurate. It is true that the first attempts to map PdfR expression with antisera and GAL4s were inaccurate. However, subsequent work by several groups has produced strong convergent evidence that with the exception of the l-LNvs after several days post-eclosion, PdfR is expressed in the Cryptochrome expressing a subset of the clock neuron network. This section of the study should be revised.

We thank the reviewer for pointing this out. As we have already addressed and revised the related part in the RESULTS section (Line 308-317), we have now removed this part from the DISCUSSION section of the revised version.

Point 3-One minor issue that would avoid unnecessary confusion by readers familiar with the circadian literature is the say that activity profiles are plotted in the study. The authors have centered their averaged activity profiles on the 12h of darkness. This is the opposite of the practice of the field, and it leads to some initial confusion in the examination of the morning and evening peak data. The authors may wish to avoid this by centering their activity plots on the 12h light phase, which would put the morning peak on the left and the evening peak on the right. This is the way the field is accustomed to examining locomotor activity profiles.

The centering of averaged activity profiles on the 12 h of darkness is done to highlight the phenotype of advanced morning activity. To prevent any confusion among readers, we have included a sentence in the figure legend explaining the difference in our activity profiles compared to previous literatures: "Activity profiles were centered of the 12 h darkness in all figures with evening activity on the left and morning activity on the right, which is different from general circadian literatures. (Fig. 2H legend)" (Line 957-959))

Point 4-The authors conclude that the loss of PDF and CNMa have opposite effects on the morning peak of locomotor activity (line 392). But they also acknowledge, briefly, that things are not that simple: loss of CNMa causes a phase advance, but loss of PDF causes a loss or reduction in the anticipatory peak. It is still significant to find a peptide transmitter with the clock neuron network that regulates morning activity, but the authors should revise their conclusion regarding the opposing actions of PDF and CNMa, which is not well supported by the data.

We have revised the relevant parts.

ABSTRACT: “Specific elimination of each from clock neurons revealed that loss of the neuropeptide CNMa in two posterior dorsal clock neurons (DN1ps) or its receptor (CNMaR) caused advanced morning activity, indicating a suppressive role of CNMa-CNMaR on morning anticipation, opposite to the promoting role of PDF-PDFR on morning anticipation.” (Line 43-48)

DISCUSSION: “Furthermore, given that the morning anticipation vanishing phenotype of Pdf or Pdfr mutant indicates a promoting role of PDF-PDFR signal, while the enhanced morning anticipation phenotype of CNMa mutant suggests an inhibiting role of CNMa signal, we consider the two signals to be antagonistic.” (Line 492-495)

Point 5-The authors should acknowledge, cite, and incorporate the substantive discussion of CNMa peptide and the DN1p neuronal class in Reinhard et al. 2022 (Front Physiol. 13: 886432).

We have revised the text accordingly and cited this paper: “Type I with two neurons whose branches projecting to the anterior region, as in CNMa∩GMR51H05, CNMa∩Pdfr, and CNMa∩GMR79A11 (Fig. 6E, 5G, 6H), and type II with four neurons branching on the posterior side with few projections to the anterior region, as in CNMa∩GMR91F02 (Fig. 6F). These two types of DN1ps’ subsets were also reported and profound discussed previously (Lamaze et al., 2018; Reinhard et al., 2022)”. (Line 393-397)

Reviewer #3 (Recommendations For The Authors):

Point 1-Throughout the manuscript figure legends (axis, genotypes, etc) are too small to be appreciated. Fig. 1. Panel A. The labels are very difficult to read.

We have attempted to enlarge the font as much as possible in the revised version.

Point 2-Fig. 1. H-J Why is efficiency not mentioned in all the examples?

In the revised manuscript, the results of Fig 1H-1J are discussed in the revised version (Line 145-147). The reason that we did not calculate the exact efficiency is that the GFP intensity is not stable enough which might change during dissection, mounting or intensity of laser in our experimental process. Therefore, in all results related to GFP signal (Fig. 1B-1J, Fig. S1, Fig. S2, Fig. 2B-2D), we relied on qualitative judgment rather than quantitative judgment, unless the GFP signal was easily quantifiable (such as in cases with limited cells or no GFP signal in the experimental group).

Point 3-Fig. 1. Panel L, left (light phase): the statistical comparisons are not clearly indicated (the same happens in Figs 3Q and 3R).

We have now re-arranged Fig. 1L and Fig. 3Q-3R to make the statistical comparisons clear in the new version.

Point 4-Line 792. Could induced be introduced?

Yes, we have now corrected this typo.

Point 5-Fig. S1. Check labels for consistency. GMR57C10 Gal4 driver is most likely R57C10.

We have now revised the labels (Fig. S1).

Point 6-Fig. S2. If the experiments were repeated and several brains were observed, the authors should include the efficiency and the number of flies as reported in Fig. S1.

We have now added the number of flies in Fig. S2 as reported in Fig. S1. As Response to Point 2 mentioned, due to the instability of the GFP signal, we are unable to provide a quantitative efficiency in this context.

Point 7-Fig S4. The fig legend describes panels I-J which are not shown in the current version of the manuscript.

We now have deleted them.

Point 8-Fig 2I. Surprising values for morning anticipation indexes even for controls (0.5 would indicate ¨no anticipation¨; in controls, the expected values would be >>0.5, as most of the activity is concentrated right before the transition. Could the authors explain this unexpected result?

We have revised the description of the calculation in the methods section (Line 612). After calculating the ratio of the last three hours of activity to the total six hours of activity, the results were further subtracted by 0.5. Therefore, the index should be ≤0.5. When the index is equal to 0, it indicates no morning anticipation.

Point 9-Fig 2K/L. The authors mention that not all genes are effectively knocked out with their strategy. Could this be accounted for the specific KD strategy, its duration, or the promotor strength? It is surprising no explanation is provided in the text (page 9 line 179).

In our pursuit of establishing a broadly effective method for gene editing, Fig. 2H-2L and Fig. 2D revealed that previous attempts have fallen short of achieving this objective. The observed inefficiency may be attributed to the intensity of the promoter, resulting in inadequate expression. Alternatively, the insufficient duration of the operation may also contribute to the lack of success. However, in the context of sleep and rhythm research applications, the age of the fruit fly tests is typically fixed, limiting the potential to enhance efficiency by extending the manipulation time. Moreover, increasing the expression level may pose challenges related to cytotoxicity, as reported in previous studies (Port et al., 2014). We refrain from offering specific explanations, as we lack a definitive plan and cannot provide additional robust evidence to support the above speculations. Consequently, in our ongoing efforts, we aim to enhance the efficiency of the tool system while operating within the current constraints.

Point 10-Page 9, line 179. Can the authors include a brief description of the reason for the different modifications? Only one was referenced.

We have revised related part in the manuscript (Line 223-231):

Cas9.M9: We fused a chromatin-modulating peptide (Ding et al., 2019), HMGN1 183 (High mobility group nucleosome binding domain 1), at the N-terminus of Cas9 and HMGB1 184 (High mobility group protein B1) at its C-terminus with GGSGP linker, termed Cas9.M9.

Cas9.M6: We also obtained a modified Cas9.M6 with HMGN1 at the N-terminus and an undefined peptide (UDP) at the C-terminus. (NOTE：UDP was gained by accident)

Cas9.M0: We replaced the STARD linker between Cas9 and NLS in Cas9.HC with GGSGP the linker (Zhao et al., 2016), termed Cas9.M0

Point 11-The authors tested the impact of KO nAChR2 across the different versions of conditional disruption (Fig 1K-L, Fig 2L, Fig 3R). It is surprising they observe a difference in daytime sleep upon knocking down with Cas9.HC (2L) but not with Cas9.M9 (3R) and the reverse is seen for night-time sleep. Could the authors provide an explanation? Efficiency is not the issue at stake, is it?

In Fig. 2K, the day sleep of flies (R57C10-GAL4/UAS-sgRNAnAChRbeta2; UAS-Cas9/+) was significantly decreased compared to flies (R57C10-GAL4/UAS-sgRNAnAChRbeta2; +/+), but not when compared to flies (R57C10-GAL4/+; UAS-Cas9/+). Our criterion for asserting a difference is that the experimental group must show a significant distinction from both control groups. Therefore, we concluded that there was no significant difference between the experimental group and the control groups in Fig. 2K.

Point 12-Fig. 4. Which of the two strategies described in A-B was employed to assemble the expression profile of CCT genes in clock neurons shown in C? This information should be part of the fig legend.

We have now revised the legend as follows: “(A-B) Schematic of intersection strategies used in Clk856 labelled clock neurons dissection, Flp-out strategy (A) and split-LexA strategy (B). The exact strategy used for each gene is annotated in Table S5.”

Point 13-Similarly, how many brains were analyzed to give rise to the table shown in C?

We have now revised the legend of Table S4 to address this concern. As indicated in: “The largest N# for each gene in Table S4 is the brain number analyzed for each gene”.

Point 14-Finally, the sentence ¨The figure is...¨ requires revision.

We have now revised it: “The exact cell number for each subset is annotated in Table S4”.

Point 15-Legend to Table S3. The authors have done an incredible job testing many gRNAs for each gene potentially relevant for communication. However, there is very little information to make the most out of it; for instance, the legend does not inform why many of the targeted genes do not appear to have been tested any further. It would be useful to the reader to discern whether despite being the 3 most efficient gRNAs, they were still not effective in targeting the gene of interest, or whether they showed off-targets, or it was simply a matter of testing the educated guesses. This information would be invaluable for the reader.

First, we designed and generated transgenic UAS-sgRNA fly lines for all these sgRNAs. We randomly selected 14 receptor genes, known for their difficulty in editing based on our experience, to assess the efficiency of our strategy, as depicted in Fig. 3M-3P, Fig. S5, and Fig. S6. We believe these results are representative and indicative of the efficiency of sgRNAs designed using our process and applied with the modified Cas9.

Secondly, we acknowledge your valid concern. While we selected sgRNAs with no predicted off-target effects through various prediction models (outlined in the Methods under C-cCCTomics sgRNA design), we did not conduct whole-genome sequencing. Consequently, we can only assert that the off-target possibility is relatively low. To address potential misleading effects arising from off-target concerns, it is essential to validate these results through mutants, RNAi, or alternative UAS-sgRNAs targeting the same gene.

Point 16-Table S4. Some of the data presented derives from observations made in 1-2 brains for a specific cluster; isn´t it too little to base a decision on whether a certain gene is (or not) expressed? It is surprising since the same CCT line was observed/analysed in more brains for other clusters. Can the authors explain the rationale?

The N# number represents the GFP positive number, and we have revised the legend of Table S4. The largest N# number denotes the total number of brains analyzed for a specific CCT line. It's possible that, due to variations in our dissection or mounting process, some clusters were only observed in 1-2 brains out of the total brains analyzed. To enhance the accuracy of intersection analysis results, we marked all positive signal records when positive subsets were found in less than 1/3 of the total analyzed brains (Table S4).

Point 17-The paragraph describing this data in the results section needs revising (lines 233-243).

We have now revised this. (Line 286-317)

Point 18-While it is customary for authors to attempt to improve the description of the activity patterns by introducing new parameters (i.e. MAPI and EAPI, lines 253-258) it would be interesting to understand the difference between the proposed method and the one already in use (which compares the same parameter, i.e., the slope (defined as ¨the slope of the best-fitting linear regression line over a period of 6 h prior to the transition¨, i.e., Lamaze et al. 2020 and many others). Is there a need to introduce yet another one?

This approach is necessary. The slope defined by Lamaze et al. utilizes data from only 2 time points, which may not accurately capture the pattern within a period before light on or off. Linear regression is not well-suited for a single fly due to the high variability in activity at each time point, making it challenging to fit the model at the individual level. The parameters we have introduced (MAPI and EAPI) in this paper are concise and can be applied at the individual level, effectively reflecting the morning or evening anticipation characteristics of each fly.

As an alternative, the activity plot of a certain fly line could be represented by an average of all flies' activity in one experiment. This would make linear regression easier to fit. However, several independent experiments are required for statistical robustness, necessitating the inclusion of hundreds of flies for each strain in a single analysis.

Point 19-In general, the legends of supplementary figures are a bit too brief. S7 and S8: it is not clear which of the two intersectional strategies were used (it would benefit whoever is interested in replicating the experiments). Legend to Fig S8 should read ¨similar to Fig S7¨.

We have now revised the legend and included “The exact strategy used for each gene is annotated in Table S5” in the legend.

Point 20-The legend in Table S6 should clearly state the genotypes examined. What does the marking in bold refer to?

We have now revised annotation of Table S6. Marking in bold refer to results out of one SD compared to control group.

Point 21-Line 314. The sentence needs revision.

We have revised these sentences.

Point 22-Line 391 (and also in the results section). The authors attempt to describe the CNMa phenotype as the opposite of pdf/pdfr mutant phenotypes. However, no morning anticipation/advanced morning anticipation are not necessarily opposite phenotypes.

We have revised related description.

ABSTRACT: “Specific elimination of each from clock neurons revealed that loss of the neuropeptide CNMa in two posterior dorsal clock neurons (DN1ps) or its receptor (CNMaR) caused advanced morning activity, indicating a suppressive role of CNMa-CNMaR on morning anticipation, opposite to the promoting role of PDF-PDFR on morning anticipation.” (Line 43-48)

DISCUSSION: “Furthermore, given that the morning anticipation vanishing phenotype of Pdf or Pdfr mutant indicates a promoting role of PDF-PDFR signal, while the enhanced morning anticipation phenotype of CNMa mutant suggests an inhibiting role of CNMa signal, we consider the two signals to be antagonistic.” (Line 492-495)

Reference

Deng, B., Li, Q., Liu, X., Cao, Y., Li, B., Qian, Y., Xu, R., Mao, R., Zhou, E., Zhang, W., et al. (2019). Chemoconnectomics: mapping chemical transmission in Drosophila. Neuron 101, 876-893.e874.

Ding, X., Seebeck, T., Feng, Y., Jiang, Y., Davis, G.D., and Chen, F. (2019). Improving CRISPR-Cas9 genome editing efficiency by fusion with chromatin-modulating peptides. Crispr j 2, 51-63.

Duhart, J.M., Herrero, A., de la Cruz, G., Ispizua, J.I., Pírez, N., and Ceriani, M.F. (2020). Circadian Structural Plasticity Drives Remodeling of E Cell Output. Curr Biol 30, 5040-5048.e5045.

Erion, R., King, A.N., Wu, G., Hogenesch, J.B., and Sehgal, A. (2016). Neural clocks and Neuropeptide F/Y regulate circadian gene expression in a peripheral metabolic tissue. eLife 5, e13552.

Fujiwara, Y., Hermann-Luibl, C., Katsura, M., Sekiguchi, M., Ida, T., Helfrich-Förster, C., and Yoshii, T. (2018). The CCHamide1 neuropeptide expressed in the anterior dorsal neuron 1 conveys a circadian signal to the ventral lateral neurons in Drosophila melanogaster. Front Physiol 9, 1276.

Goda, T., Tang, X., Umezaki, Y., Chu, M.L., Kunst, M., Nitabach, M.N.N., and Hamada, F.N. (2016). Drosophila DH31 neuropeptide and PDF receptor regulate night-onset temperature preference. J Neurosci 36, 11739-11754.

Goda, T., Umezaki, Y., Alwattari, F., Seo, H.W., and Hamada, F.N. (2019). Neuropeptides PDF and DH31 hierarchically regulate free-running rhythmicity in Drosophila circadian locomotor activity. Sci Rep 9, 838.

Guo, F., Cerullo, I., Chen, X., and Rosbash, M. (2014). PDF neuron firing phase-shifts key circadian activity neurons in Drosophila. Elife 3.

He, C., Cong, X., Zhang, R., Wu, D., An, C., and Zhao, Z. (2013). Regulation of circadian locomotor rhythm by neuropeptide Y-like system in Drosophila melanogaster. Insect Mol Biol 22, 376-388.

Hermann, C., Yoshii, T., Dusik, V., and Helfrich-Förster, C. (2012). Neuropeptide F immunoreactive clock neurons modify evening locomotor activity and free-running period in Drosophila melanogaster. J Comp Neurol 520, 970-987.

Hyun, S., Lee, Y., Hong, S.T., Bang, S., Paik, D., Kang, J., Shin, J., Lee, J., Jeon, K., Hwang, S., et al. (2005). Drosophila GPCR Han is a receptor for the circadian clock neuropeptide PDF. Neuron 48, 267-278.

Johard, H.A., Yoishii, T., Dircksen, H., Cusumano, P., Rouyer, F., Helfrich-Förster, C., and Nässel, D.R. (2009). Peptidergic clock neurons in Drosophila: ion transport peptide and short neuropeptide F in subsets of dorsal and ventral lateral neurons. J Comp Neurol 516, 59-73.

Lamaze, A., Krätschmer, P., Chen, K.F., Lowe, S., and Jepson, J.E.C. (2018). A Wake-Promoting Circadian Output Circuit in Drosophila. Curr Biol 28, 3098-3105.e3093.

Lear, B.C., Zhang, L., and Allada, R. (2009). The neuropeptide PDF acts directly on evening pacemaker neurons to regulate multiple features of circadian behavior. PLoS Biol 7, e1000154.

Lee, G., Bahn, J.H., and Park, J.H. (2006). Sex- and clock-controlled expression of the neuropeptide F gene in Drosophila. 103, 12580-12585.

Lelito, K.R., and Shafer, O.T. (2012). Reciprocal cholinergic and GABAergic modulation of the small ventrolateral pacemaker neurons of Drosophila's circadian clock neuron network. J Neurophysiol 107, 2096-2108.

Ma, D., Przybylski, D., Abruzzi, K.C., Schlichting, M., Li, Q., Long, X., and Rosbash, M. (2021). A transcriptomic taxonomy of Drosophila circadian neurons around the clock. Elife 10.

Port, F., Chen, H.M., Lee, T., and Bullock, S.L. (2014). Optimized CRISPR/Cas tools for efficient germline and somatic genome engineering in Drosophila. Proc Natl Acad Sci USA 111, E2967-2976.

Reinhard, N., Schubert, F.K., Bertolini, E., Hagedorn, N., Manoli, G., Sekiguchi, M., Yoshii, T., Rieger, D., and Helfrich-Förster, C. (2022). The Neuronal Circuit of the Dorsal Circadian Clock Neurons in Drosophila melanogaster. Front Physiol 13, 886432.

Renn, S.C., Park, J.H., Rosbash, M., Hall, J.C., and Taghert, P.H. (1999). A pdf neuropeptide gene mutation and ablation of PDF neurons each cause severe abnormalities of behavioral circadian rhythms in Drosophila. Cell 99, 791-802.

Shafer, O.T., Helfrich-Förster, C., Renn, S.C., and Taghert, P.H. (2006). Reevaluation of Drosophila melanogaster's neuronal circadian pacemakers reveals new neuronal classes. J Comp Neurol 498, 180-193.

Shafer, O.T., Kim, D.J., Dunbar-Yaffe, R., Nikolaev, V.O., Lohse, M.J., and Taghert, P.H. (2008). Widespread receptivity to neuropeptide PDF throughout the neuronal circadian clock network of Drosophila revealed by real-time cyclic AMP imaging. Neuron 58, 223-237.

Zhang, L., Chung, B.Y., Lear, B.C., Kilman, V.L., Liu, Y., Mahesh, G., Meissner, R.A., Hardin, P.E., and Allada, R. (2010). DN1(p) circadian neurons coordinate acute light and PDF inputs to produce robust daily behavior in Drosophila. Curr Biol 20, 591-599.

Zhao, P., Zhang, Z., Lv, X., Zhao, X., Suehiro, Y., Jiang, Y., Wang, X., Mitani, S., Gong, H., and Xue, D. (2016). One-step homozygosity in precise gene editing by an improved CRISPR/Cas9 system. Cell Res 26, 633-636.

Version 2

Abstract

Abstract

Dissection of neural circuitry underlying behaviors is a central theme in neurobiology. We have previously proposed the concept of chemoconnectome (CCT) to cover the entire chemical transmission between neurons and target cells in an organize and created tools for studying it (CCTomics) by targeting all genes related to the CCT in Drosophila. Here we have created lines targeting the CCT in conditional manners after modifying GFP RNA interference, Flp-out and CRISPR/Cas9 technologies. All three strategies are validated to be highly effective with the best using chromatin-peptide fused Cas9 variants and scaffold optimized sgRNAs. As a proof of principle, we conduct a comprehensive intersection analysis of CCT genes expression profiles in the clock neurons, uncovering 43 CCT genes present in clock neurons. Specific elimination of each from clock neurons revealed that loss of the neuropeptide CNMa in two posterior dorsal clock neurons (DN1ps) or its receptor (CNMaR) caused advanced morning activity, indicating a suppressive role of CNMa-CNMaR on morning anticipation, opposite to the promoting role of PDF-PDFR on morning anticipation. These results demonstrate the effectiveness of conditional CCTomics and its tools created by us here and establish an antagonistic relationship between CNMa-CNMaR and PDF-PDFR signaling in regulating morning anticipation.

Introduction

Much research efforts have been made to uncover the wiring and signaling pathways of neural circuits underlying specific behaviors. Circuit dissection strategies include genetic screening, genetic labelling, circuit tracing, live imaging, genetic sensors and central nervous system (CNS) reconstruction via electron microscopy (EM). Recently, we have developed chemoconnectomics (CCTomics), focusing on building a comprehensive set of knockout and knockin tool lines of chemoconnectome (CCT) genes, to dissect neural circuitry based on chemical transmission (Deng et al., 2019).

Each strategy has advantages and disadvantages. For example, genetic screening is less biased but inefficient; circuit tracing with viruses provides information of connection, but is often prone to leaky expression and inaccurate labelling; and EM reconstruction is anatomically accurate but does not allow for manipulation of corresponding neurons. CCTomics overcomes limitations of previous strategies by allowing for behavioral screening of CCT genes and accurate labelling or manipulation of corresponding neurons. However, it is still limited in that knockout of some CCT genes can be lethal during development and that CCT genes may function differently in different neurons, which require a cell type-specific manipulation. Thus, we decided to invent a conditional CCTomics (cCCTomics) in which gene deletion was conditional.

There are three major strategies for somatic gene mutagenesis at the DNA/RNA level: RNA interference, DNA site-specific recombination enzymes, and CRISPR/Cas system. RNA interference targets RNAs conveniently and efficiently (Martin and Caplen, 2007; Oberdoerffer et al., 2005). Libraries of transgenic RNAi flies covering almost the entire fly genome have been established (Ni et al., 2011; Perkins et al., 2015). DNA site specific recombination enzymes such as Flp, B3 and Cre mediate specific and efficient gene editing (Gaj et al., 2014; Grindley et al., 2006). These strategies require flies with reverse repetitive sequences knocked into the corresponding genes, which is time-consuming with relatively complex recombination for genetic assays. CRISPR/Cas systems, particularly CRISPR/Cas9, which targets DNA with a sgRNA/Cas protein complex, have been broadly applied in gene manipulation over the last decade. The widespread use of CRIPSR/Cas9 in Drosophila somatic gene manipulation began in 2014 (Xue et al., 2014). Later, tRNA-flanking sgRNAs was designed and applied, which enabled multiple sgRNAs to mature in a single transcript (Xie et al., 2015), accelerating the application of this strategy in conditional gene manipulation in flies with impressive efficiency (Delventhal et al., 2019; Port and Bullock, 2016; Port et al., 2020; Schlichting et al., 2019; Schlichting et al., 2022). Additionally, libraries of UAS-sgRNA targeting kinases (Port et al., 2020) and GPCRs (Schlichting et al., 2022) have been established, but no sgRNA libraries covering all the CCT genes exists yet. The efficiency of CRISPR/Cas9 has not been validated systematically in the Drosophila nervous system.

The circadian rhythm can be used for proof of principle testing of cCCTomics. Organisms evolve periodic behaviors and physiological traits in response to cyclical environmental changes. The rhythmic locomotor behavior of Drosophila, for instance, shows enhanced activity before the light is turned on and off in a light-dark (LD) cycle, referred to as morning and evening anticipations, respectively (Collins et al., 2005; Helfrich-Förster, 2001). Under 12 hours (h) dark-12 h dark (DD) conditions, the activities peaks regularly about every 24 h (Konopka and Benzer, 1971). Approximately 150 clock neurons, circadian output neurons and extra-clock electrical oscillators (xCEOs) coordinate Drosophila circadian behaviors (Dubowy and Sehgal, 2017; Tang et al., 2022). The regulation of morning and evening anticipations, the most prominent features in the LD condition, is primarily mediated by four pairs of sLNvs expressing pigment dispersing factor (PDF), six pairs of LNds and the 5th s-LNv (Grima et al., 2004; Rieger et al., 2006; Stoleru et al., 2004). At the molecular level, Pdf and Pdf receptor (PDFR) are well-known, with their mutants showing an advanced evening activity peak and no morning anticipation (Hyun et al., 2005; Lear et al., 2005; Renn et al., 1999). Other neuropeptides and their receptors, including AstC/AstC-R2 and neuropeptide F (NPF) and its receptor (NPFR), have also been reported to modulate evening activities (Díaz et al., 2019; He et al., 2013; Hermann et al., 2012), while CCHa1/CCHa1-R and Dh31 regulate morning activities (Fujiwara et al., 2018; Goda et al., 2019). To date, no advanced morning activity phenotype has been reported in flies.

To develop a more efficient approach for somatic gene manipulation, we have now generated two systems for conditional manipulation of CCT genes. One is GFPi/Flp-out-based conditional knockout system of CCT genes (cCCTomics) and another is CRISPR/Cas9-based (C-cCCTomics). Both systems have achieved high efficiency of gene mutagenesis in the Drosophila nervous system. C-cCCTomics, utilizing chromatin-peptide fused Cas9 and scaffold optimized sgRNA, makes efficient conditional gene knockout as simple as RNAi. Further application of C-cCCTomics in clock neurons revealed novel roles of CCT genes in circadian behavior: CNMa-CNMaR modulates morning anticipation as an antagonistic signal of PDF-PDFR.

Results

Near Complete Disruption of Target Genes by GFPi and Flp-out Based cCCTomics

For the purpose of conditional chemoconnectomics (cCCTomics), we initially leveraged the benefits of our previously generated CCTomics attP lines (Deng et al., 2019), which enabled us to fuse enhanced GFP coding sequence at the 3’ end of each gene’s coding region and flank most or entire gene span with FRT sequence through site-specific recombination (Fig. 1A, Table S1). We designed this system so that it could be used to target genes tagged with GFP by RNAi (Neumüller et al., 2012) and at the same time to enable flippase(Flp) mediated DNA fragment excision between two FRT sequences when FRT sequences are in the same orientation (Vetter et al., 1983) (Fig. 1A).

cCCTomics Mediates Efficient Conditional Disruption of CCT Genes

(A) Schematic of cCCT gene span and principle of cCCTomics. A T2A-EGFP sequence was introduced at the 3’ end of CCT genes and their most or all coding regions (depending on attP-KO lines) were flanked by 34 bp FRT sequence. Both Flp-out (top) and GFP RNAi (down) could mediate CCT gene manipulation. (B-J) Expression of Trh (B-D), CCHa2 (E-G), and Dh31 (H-J) are efficiently disrupted by pan-neuronal expression of GFP-RNAi (B-D), pan-neuronal expression of Flp-out (E-G), and heatshock-Flp (H-J) respectively. Representative fluorescence images of R57C10-Gal4/+;TrhEGFP.FRT/+ (B), UAS-shRNAGFP/TrhEGFP.FRT (C), R57C10-Gal4/+; UAS-shRNAGFP/TrhEGFP.FRT (D), R57C10-Gal4/+;CCHa2EGFP.FRT/+; (E), UAS-Flp/CCHa2EGFP.FRT (F), R57C10-Gal4/+; UAS-Flp/CCHa2EGFP.FRT (G), Dh31EGFP.FRT with heatshock (H), hs-Flp/Dh31EGFP.FRT without heatshock (I), and hs-Flp/Dh31EGFP.FRT with heatshock are shown. Manipulation efficiency and experiment group fly number is noted on the right. Scale bar, 50um. (K-L) sleep profiles (K) and statistical analysis (L) of Flp-out induced nAChRβ2 neuronal knockout flies (dark red), nAChRβ2 knockout flies, and genotype controls (dark and blue). Sleep profiles are plotted in 30 min bins. In this and other figures, blank background indicates the light phase (ZT 0-12); shaded background indicates the dark phase (ZT 12-24). Daily sleep duration were significantly reduced in nAChRβ2 neuronal knockout files which is comparable to nAChRβ2 knockout. In all statistical panels, unless otherwise noted, 1) Numbers below each bar represent the number of flies tested. 2) Mean ± SEM is shown. 3) The Kruskal-Wallis test followed by Dunn’s post test was used. ***p<0.001. ** p<0.01. * p<0.05. n.s. p>0.05. Male flies were used unless otherwise noted.

To validate the efficiency of cCCTomics, we performed pan-neuronal expression of either shRNAGFP or flipase in cCCT flies. Immunofluorescent imaging showed that constitutive expression of shRNAGFP (Fig. 1B-1D, Fig. S1A-S1I) or flipase (Fig. 1E-1G, Fig. S1J-S1U) almost completely eliminated GFP signals of target genes, indicating high efficiency. Knocking out at the adult stage using either hsFLP driven Flp-out (Golic and Lindquist, 1989) (Fig. 1H-1J) or neural (elav-Switch) driven shRNAGFP (Nicholson et al., 2008; Osterwalder et al., 2001) (Fig. S2A-S2I), also resulted in the elimination of most, though not all, GFP signals. Notably, control group of CCTEGFP.FRT; elav-Switch/UAS-shRNAGFP flies fed with solvent (ethanol) showed obvious decreased GFP (Fig. S2B, S2E, and S2H) comparing with UAS-shRNAGFP/CCTEGFP.FRT flies fed with RU486 (Fig. S2A, S2D, and S2G), indicating leaky expression of elav-Switch.

We then applied cCCTomics in pan-neuronal knockout of nAChRβ2 which is required for Drosophila sleep (Dai et al., 2021). Ablation of nAChRβ2 in the nervous system dramatically decreased sleep of flies, mirroring the nAChRβ2 knockout phenotype (Fig. 1K, 1L). Therefore, cCCTomics is an effective toolkit for manipulation of CCT genes and suitable for functional investigations of genes. Expression of in-frame fused eGFP-labelled CCT genes highly co-localized with signals revealed by immunocytochemistry (Fig. S3A-S3I), allowing direct examination of gene expression without amplification, which is different from the GAL4/UAS binary system.

We then checked the viability of cCCT lines and found that cCCT lines including CapaEGFP.FRT, ChATEGFP.FRT and EhEGFP.FRT were viable whereas their CCT mutants were lethal. Gad1EGFP.FRT, GluRIIDEGFP.FRT and CapaREGFP.FRT were still lethal as their CCT mutants were (Table S2). This indicates that some of the cCCT knockin flies may functionally affect corresponding genes, which are not suitable for conditional gene manipulation. Combination of cCCT transgenic flies with UAS-flp, UAS-shRNAGFP or specific drivers is complicated and unfriendly for screen work, despite the almost 100% efficiency of gene suppression when driven by a pan-neuronal driver. Because of the limitations of this method, we further created a CRISPR/Cas9 based conditional knockout system of chemoconnectomics (C-cCCTomics).

CRISPR/Cas9-Based Conditional Knockout System for CCTomics

To simplify effective manipulation of CCT genes, we designed a vector based on pACU2 (Han et al., 2011) with tRNA flanking sgRNAs (Port and Bullock, 2016; Xie et al., 2015) targeting CCT genes (Fig. 2A). We also adopted an optimized sgRNA scaffold “E+F” (E, stem extension; F, A-U flip) (Chen et al., 2013), which facilitates Cas9-sgRNA complex formation and gene knockout efficiency (Dang et al., 2015; Poe et al., 2019; Zhao et al., 2016), to all sgRNAs to improve gene knockout efficiency. To balance efficiency and off-target effect, we selected three sgRNAs for each CCT gene with highest predicted efficiency and no predicted off-target effect based on previously reported models (Chu et al., 2016; Doench et al., 2014; Graf et al., 2019; Gratz et al., 2014; Heigwer et al., 2014; Stemmer et al., 2015; Xu et al., 2015) (see Table S3 and details at Methods).

C-cCCTomics Mediates Efficient Conditional Knockout of CCT Genes

(A) Schematic of C-cCCTomics principle. Cas9 and three sgRNAs are driven by GAL4/UAS system. Three tandem sgRNAs are segregated by fly tRNAGly and matured by RNase Z and RNase P. (B-G) Pan-neuronal knockout of Dh31 (D) and Pdf (G) by C-cCCTomics strategy. Representative fluorescence images presented expression of Dh31 (B-D) or anti-PDF (E-G). Pan-neuronal expression of Cas9 and sgRNA eliminated most (D) or all (G) fluorescent signal compared to control fly brains (B-C, E-F). Scale bar, 50 μm. (H) Activity profiles of pan-neuronal knockout of Pdfr and Pdfr-attpKO. Activity profiles were centered of the 12h darkness in all figures with evening activity on the left and morning activity on the right, which is different from general circadian literatures. Plotted in 30 min bins. (I) Statistical analysis of morning anticipation index (MAI), power, and period for pan-neuronal Pdf knockout and Pdfr-attpKO flies. Knocking out of Pdfr in neurons reduced both morning anticipation index, power and period significantly. (J-K) Statistical analysis of nAChRα2 (J) and nAChRβ2 (K) pan-neuronal knockout flies’ sleep phenotype. Sleep of these flies were not disrupted.

We generated UAS-sgRNA3x transgenic lines for all 209 defined CCT genes (Abruzzi et al., 2017; Dai et al., 2019; Deng et al., 2019) and UAS-Cas9.HC (Cas9.HC) (Mali et al., 2013). We first verified that C-cCCTomics mediated precise target DNA breaking by ubiquitous expression of Cas9.P2 (Port et al., 2014) and sgRNA by targeting Pdf or Dh31. Sanger sequencing showed that indels were present exactly at the Cas9 cleavage sites (Fig. S4A-S4H).

To determine the efficiency of C-cCCTomics, we employed pan-neuronal expression of Cas9.HC with sgRNADh31 or sgRNApdf. Targeting by Cas9.HC/sgRNADh31 eliminated most but not all of the GFP signal in Dh31EGFP.FRT (Fig 2B-2D), whereas all anti-PDF signals were eliminated by Cas9.HC/sgRNApdf (Fig 2E-2G). Furthermore, we used the C-cCCT strategy to conditionally knockout genes for pdfr, nAChRβ2 and nAChRα2, which were previously reported as essential for circadian rhythm or sleep (Dai et al., 2021; Renn et al., 1999). Pan-neuronal knockout of Pdfr resulted in a tendency towards advanced evening activity and weaker morning anticipation compared to control flies (Fig. 2H-2I), which is similar to Pdfr-attpKO flies. These phenotypes were not as pronounced as those reported previously, when han5304 mutants exhibited a more obvious advanced evening peak and no morning anticipation (Hyun et al., 2005). Furthermore, there was no significant sleep decrease in these conditional knockout (cKO) flies (Fig. 2J-2K) when we applied C-cCCTomics to manipulate nAChRβ2 or nAChRa2. Taken together, C-cCCTomics (with Cas9.HC) achieved a relatively high gene knockout efficiency, but it was not effective enough for all genes.

Evaluation of Cas9 with Different Chromatin-Modulating Peptides

Since the establishment of the CRISPR/Cas9 system a decade ago, many groups have attempted to improve its efficiency in gene manipulation. Most attempts have been focused on the two main components of this system, the Cas9 protein (Ding et al., 2019; Ling et al., 2020; Liu et al., 2019; Zhao et al., 2016; Zheng et al., 2020) and the single guide RNA (Chen et al., 2013; Chu et al., 2016; Dang et al., 2015; Doench et al., 2014; Filippova et al., 2019; Graf et al., 2019; Labuhn et al., 2018; Mu et al., 2019; Nahar et al., 2018; Scott et al., 2019; Xu et al., 2015). At the beginning of C-cCCTomics design, we adopted an optimized sgRNA scaffold and selected sgRNAs with predicted high efficiency. We tried to further improve the efficacy by modifying Cas9 protein. We fused a chromatin-modulating peptide (Ding et al., 2019), HMGN1 (High mobility group nucleosome binding domain 1), at the N-terminus of Cas9 and HMGB1 (High mobility group protein B1) at its C-terminus with GGSGP linker, termed Cas9.M9 (Fig. 3A, Methods). We also obtained a modified Cas9.M6 with HMGN1 at the N-terminus and an undefined peptide (UDP) at the C-terminus (Fig. 3A). We replaced the STARD linker between Cas9 and NLS in Cas9.HC with GGSGP the linker (Zhao et al., 2016), termed Cas9.M0 (Fig 3A). None of these modifications have been validated previously in flies.

Efficiency Evaluation of Variations of Chromatin-Modulating Peptides Modified Cas9.

(A) Schematics of chromatin-modulating peptides modified Cas9. (B-J) Efficiency evaluation of Cas9 variants. Fluorescence imaging of R57C10-Gal4>UAS-sgRNAple (B), R57C10-Gal4>UAS-Cas9 (C-F), and R57C10-Gal4>UAS-Cas9, UAS-sgRNAple (G-J) flies are shown. Brains were stained with anti-TH (green). Scale bar is 50μm. (K) Anterior TH positive neurons numbers of (K-U). (L) Statistical analysis of ple knockout efficiency related to (K). Modified Cas9.M6 and Cas9.M9 showed an improved efficiency comparing to Cas9.HC. Student’s t test was used. (M) Statistical analysis of representational KO efficiency of Cas9 variants as related to Figure S5. Gene symbols on the right indicate tested genes. (N) Statistical analysis of Cas9 expression level. (O-P) Workflow of efficiency validation by next-generation sequencing (O) and Statistical analysis of single-site mutation ratios induced by Cas9 variants (P). Paired t test was used in (M), (N) and (P). (Q-R) Statistical analysis of sleep amount for nAChRα2 (Q) or nAChRβ2 (R) pan-neuronal knockout flies. Knockout of nAChRα2 and nAChRβ2 by modified Cas9.M9 significantly decreased flies’ sleep amount.

To determine whether the modified Cas9 variants were more efficient, we first pan-neuronally expressed each Cas9 variant and sgRNAple, and assessed their efficiency by immunofluorescence imaging. By counting anti-TH positive neurons in the brain (anterior view) after targeting by Cas9/sgRNAple, we found that unmodified Cas9.HC/sgRNAple only achieved 69.58±3.04% (n=5) knockout efficiency (Fig. 3G, 3K, 3L), while Cas9.M6/sgRNAple and Cas9.M9/sgRNAple significantly improved efficiency to 87.53±3.06% (n=7) and 97.19±2.15% (n=8), respectively (Fig. 3I-3L). Fourteen additional CCT genes were subjected to pan-neuronal knockout, and the mRNA levels of the target genes were evaluated using real-time quantitative PCR with at least one primer overlapping the sgRNA targeting site (Fig. S5). Cas9.M6 and Cas9.M9 demonstrated significantly higher gene disruption efficiency compared to the unmodified Cas9.HC, achieving average efficiencies of 87.51%±2.24% and 89.59%±2.39% for Cas9.M6 and Cas9.M9, respectively, in contrast to 70.72%±3.82% for Cas9.HC. (Fig. 3M, Fig. S5). To rule out the possibility of the observed variations in gene disruption efficiency being attributed to differential Cas9 expression levels, we quantified the Cas9 expression levels and noted that both Cas9.M6 and Cas9.M9 exhibited lower mRNA levels than Cas9.HC under the experiment condition (Fig. 3N). Subsequently, genomic DNA of Drosophila head was extracted, and libraries encompassing target sites were constructed for high-throughput sequencing to verify disparities in genetic editing efficiency among these three Cas9 variants (Fig. 3O). In almost all nineteen sites tested, the mutation ratio consistently showed a trend towards Cas9.M6 and Cas9.M9 having a higher gene disruption efficiency than Cas9.HC (Fig. 3P, Fig. S6). The single-site mutation rates varied from 5.81% to 43.47% for Cas9.HC, 22.40% to 53.54% for Cas9.M6, and 19.90% to 63.57% for Cas9.M9 (Fig. 3P, Fig. S6). It should be noted that genomic DNA extracted from fly heads contained glial cells, which did not express Cas9/sgRNA, leading to a larger denominator and consequently reducing the observed mutation rates. Unmodified Cas9 displayed mutation rates comparable to those previously reported by Schlichting et al., 2022. The findings indicated that both Cas9.M6 and Cas9.M9 displayed elevated efficiency compared to Cas9.HC, with Cas9.M9 exhibiting the highest mutagenesis proficiency. These results suggest that the implementation of modified C-cCCTomics using Cas9.M6 and Cas9.M9 achieved an elevated level of efficiency. While unmodified C-cCCTomics was not efficient enough to knockout nAChRβ2 and nAChRα2 to phenocopy their mutants, we employed Cas9.M9 in conditional knockout of these genes to verify its efficiency. Pan-neuronal knocking out of nAChRβ2 or nAChRa2 by Cas9.M9/sgRNA showed significant sleep decrease which was similar to their mutants (Fig. 3Q-3R) (Dai et al., 2021).

Taken together, our results support that we have created a high efficiency toolkit for CCT gene manipulation in the nervous system, as well as more efficient Cas9 variants, Cas9.M6 and Cas9.M9, which can also be applied to genes other than those in the CCT.

CCT Genes Detected in Clock Neurons by Genetic Intersection

We analyzed the expression profile of CCT genes in circadian neurons with CCTomics driver lines in all clock neurons expressing Clk856 (Gummadova et al., 2009). With the Flp-out or split-lexA intersection strategy (Fig. 4A, 4B), we found 43 out of 148 analyzed CCT genes expressed in circadian neurons (Fig. 4C, Fig S7-S8, Table S4, Table S5). In all eight subsets of clock neurons, 23 CCT genes were expressed in DN1s, 20 in DN2s, 22 in DN3s, 28 in LNds, 14 in l-LNvs, 12 in s-LNvs, 5 in 5th s-LNv and 3 in LPNs, with a total of 127 gene-subsets.

Genetic Dissection of Clk856 Labelled Clock Neurons.

(A-B) Schematic of intersection strategies used in Clk856 labelled clock neurons dissection, Flp-out strategy (A) and split-LexA strategy (B). The exact strategy used for each gene is annotated in Table S5. (C) Expression profiles of CCT genes in clock neurons. Gradient color denotes proportion of neurons that were positive for the CCT gene within each subset. The exact cell number for each subset is annotated in Table S4.

To assess the accuracy of expression profiles using CCT drivers, we compared our dissection results with previous reports. Initially, we confirmed the expression of CCHa1 in two DN1s (Fujiwara et al., 2018), sNFP in four s-LNvs and two LNds (Johard et al., 2009), and Trissin in two LNds (Ma et al., 2021), aligning with previous findings. Additionally, we identified the expression of nAChRα1, nAChRα2, nAChRβ2, GABA-B-R2, CCHa1-R, and Dh31-R in all or subsets of LNvs, consistent with suggestions from studies using ligands or agonists in LNvs (Duhart et al., 2020; Fujiwara et al., 2018; Lelito and Shafer, 2012; Shafer et al., 2008) (Table S4).

Regarding previously reported Nplp1 in two DN1as (Shafer et al., 2006), we found approximately five DN1s positive for Nplp-KI-LexA, indicating a broader expression than previously reported. A similar pattern emerged in our analysis of Dh31-KI-LexA, where four DN1s, four s-LNvs, and two LNds were identified, contrasting with the two DN1s found in immunocytochemical analysis (Goda et al., 2016). Colocalization analysis of Dh31-KI-LexA and anti-PDF revealed labeling of all PDF-positive s-LNvs but not l-LNvs (Fig. S9A), suggesting that the differences may arise from the broader labeling of 3’ end knock-in LexA drivers or the amplitude effect of the binary expression system. The low protein levels might go undetected in immunocytochemical analysis. This aligns with transcriptome analysis findings showing Nplp1 positive in DN1as, a cluster of CNMa-positive DN1ps, and a cluster of DN3s(Ma et al., 2021), which is more consistent with our dissection.

Despite the well-known expression of PDF in LNvs and PDFR in s-LNvs (Shafer et al., 2008), we did not observe stable positive signals for both in Flp-out intersection experiments, although both Pdf-KI-LexA and Pdfr-KI-LexA label LNvs as expected (Fig. S9B-S9C). We also noted fewer positive neurons in certain clock neuron subsets compared to previous reports, such as NPF in three LNds and some LNvs (Erion et al., 2016; He et al., 2013; Hermann et al., 2012; Johard et al., 2009; Lee et al., 2006) and ChAT in four LNds and the 5th s-LNv (Duhart et al., 2020; Johard et al., 2009) (Table S4). We attribute this limitation to the inefficiency of LexAop-FRT-myr::GFP driven by LexA, acknowledging that our intersection results may miss some positive signals.

Conditional Knockout of CCT Genes in Clock Neurons

To investigate the function of CCT genes in circadian neurons with our conditional knockout system, we knocked out all 67 (genes identified above and reported previously) CCT genes in Clk856-labeled clock neurons by C-cCCTomics.

In the pilot screen, we monitored fly activity by video recording (Dai et al., 2019) and analyzed rhythmic behavior under LD and DD conditions. We analyzed morning anticipation index (MAI) and evening anticipation index (EAI) under the LD condition (Harrisingh et al., 2007; Im and Taghert, 2010; Seluzicki et al., 2014) (Fig. 5A), power, period and arrhythmic rate (AR) under the DD condition. Fly activities tended to rise rapidly after ZT22.5 at dawn and ZT10 at dusk. Thus, we added two more parameters to describe the anticipatory activity patterns of LD condition. Morning anticipation pattern index (MAPI) was defined as the difference between Pia[arctan(ZT20.5∼ZT22.5 activity increasing slope)] and Pip[arctan(ZT22.5 ∼ZT24 activity increasing slope)], M(Pia-Pip). Evening anticipation pattern index (EAPI) was defined similar to MAPI (Fig. 5A, see Methods). Pia and Pip were positive, while MAPI and EAPI were negative, for wild type (wt) flies as their activities gradually increases at dawn or dusk at increasing rates.

CNMa Regulation of Morning Anticipation in Clock Neuron

(A)Schematic of MAI, EAI, MAPI and EAPI definition. (B) Activity plots of male flies with CNMa knockout in clock neurons (red) and controls (blue, black and grey), plotted in 30 min bins. An advancement of morning activity peak was presented in CNMa clock-neuron-specific mutants (brown arrowhead). (C-E) Statistical analyses of MAPI, power, and period of flies in (B). MAPI was significantly increased in clock neurons-specific CNMa deficient flies (C) while power (D) and period (E) were not changed. (F) Schematic of CNMaKO generation. The entire encoding region of CNMa was replaced by an attP-SAstop-3P3-RFP-loxP cassette using CRISPR-Cas9 strategy. (G) Statistical analysis of MAPI of male CNMaKO flies (red) and controls (blue and black). MAPI significantly increased in male CNMaKO flies. (H) Activity plots of female CNMaKO flies (red) and controls (blue and black). (I) Statistical analysis of MAPI of female CNMaKO flies (red) and controls (blue and black). MAPI was significantly increased in female CNMaKO flies.

Knocking out Pdf or Pdfr in clock neurons phenocopied their mutants with lower MAI, advanced evening activity, low power, high arrhythmic rate and shorter period (Hyun et al., 2005; Lear et al., 2005; Renn et al., 1999) (Table S6, Fig S10A-S10D). The MAI-decreasing phenotype of Dh31 knockout was also reproduced in this pilot screen (Goda et al., 2019) (Table S6). All the above results verified the effectiveness of C-cCCTomics. Unexpectedly, additional experimental replications with full controls using Cas9.M9 revealed that leaky expression of Cas9.M9 and sgRNA might have caused disruption of Dh31, Dh44, Pdf and Pdfr (Fig. S10A-S10D) (see DISCUSSION), which was not suitable for neuronal specific mutagenesis of some genes. Therefore, in the following work we primarily focused on Cas9.M6 instead.

Analysis of the newly defined parameters MAPI and EAPI showed that control flies (Clk856-GLA4>UAS-Cas9.M9) had negative EAPI but slightly positive MAPI. The positive MAPI of control flies in this screen might be caused by Cas9.M9 toxicity. Only the Pdf and Pdfr clock-neuron knockout flies showed positive EAPIs, indicating an advanced evening activity (Table S6, Fig S10A, S10C). nAChRα1, MsR1, mAChR-B, and CNMa cKO flies had highest MAPI values (Table S6). We further confirmed their phenotypes using Cas9.M6 which revealed that CNMa plays a role in regulating morning anticipatory activity (Fig. S11A).

Regulation of Morning Anticipation by CNMa-positive DN1p Neurons

Conditionally knocking out CNMa in clock neurons advanced morning activity (Fig. 5B, Fig. S11B) and increased MAPI (Fig. 5C, Fig S11A, S11C), leaving the power and period intact in male flies (Fig. 5D, 5E). The same advanced morning activity phenotype were also observed in female flies (Fig. S11D-S11G). To further validate this phenotype, we generated a CNMa knockout (CNMaKO) line by replacing its whole coding region with an attP-splicing adaptor element (Deng et al., 2019) (Fig. 5F). Both male and female CNMaKO flies exhibited the same phenotypes as seen in the CNMa cKO (Fig. 5G-5I).

Previous studies have found CNMa expression in DN1 neurons (Abruzzi et al., 2017; Jin et al., 2021; Ma et al., 2021). Our intersection showed four DN1p and one DN3 CNMa positive neurons in Clk856 labelled neurons (Fig. S8 - 16, Fig. 4C, Table S4). Analysis with an endogenous CNMa-KI-GAL4 knockin driver showed that six pairs of CNMa neurons located in the DN1p region and three pairs located in the subesophageal ganglion (SOG) had the brightest GFP signals (Fig. 6A). The anatomical features of CNMa neurons were further confirmed using stingerRed and more neurons were found in regions, the anterior ventrolateral protocerebrum (AVLP), and the antennal mechanosensory and motor center (AMMC) (Fig. S12A). Dendrites of CNMa neurons were concentrated in DN1p and SOG, with their axons distributed around DN1p region, lateral horn (LH), and prow region (PRW) (Fig. 6B, 6C). Using the Trans-tango strategy (Talay et al., 2017), we also found that downstream of CNMa neurons were about fifteen pairs of neurons in the SOG, five pairs of LNd neurons, one pair of DN3 neurons and six intercerebralis (PI) neurons (Fig. 6D, arrowhead).

Expression, Projection and Trans-projection Feature of CNMa Neurons and Its Functional Subset

(A-C) Expression and projection patterns of CNMa-KI-Gal4 in the brain. Membrane, dendrites, and axon projections are labelled by mCD8::GFP (A), Denmark (B), and syt::eGFP (C) respectively. (D) Downstream neurons labelled through trans-tango driven by CNMa-KI-GAL4. Arrowheads indicate candidate downstream neurons: six neurons in PI, one pair in DN3, five pairs in LNd and about 15 pairs in SOG. (E-H) Intersection of DN1p CNMa neurons with DN1p labelled drivers. GMR51H05-GAL4 (E), GMR91F02-GAL4 (F), Pdfr-KI-GAL4 (G) and GMR79A11-GAL4 (H) were intersected with CNMa-p65AD, UAS-LexADBD, LexAop-myr::GFP. Two type I (E, G, H) neurons projected to anterior region and four type II (F) neurons had fewer projections to anterior region. Scale bar, 50μm. (I) MAPIs were significantly increased in all three DN1p drivers mediated CNMa knockout. (J-K) Activity plots of CNMa knockout in R51H05-GAL4 (J) and R91F02-GAL4 (N) neurons. R51H05-GAL4 mediated CNMa knockout flies showed an advanced morning activity peak (J), while R91F02-Gal4 mediated CNMa knockout flies did not (N). (Q-S) Statistical analyses of MAI, power and Period. Pdfr reintroduction in R79A11 and R51H05 neurons could partially rescue the MAI-decreased phenotype of Pdfr knockout flies. (T) Statistical analyses of MAI of Pdfr knocking out in R79A11-GAL4, CNMa-GAL4 and R51H05-GAL4 labelled neurons.

Because we had found that knocking out CNMa in Clk856-GAL4 labeled neurons produced advanced morning activity, and that CNMa intersected with Clk856-Gal4 labeled neurons in 4 pairs of DN1ps and one pair of DN3 neurons (Fig. S8-16), we focused on these neurons and performed more intersections. Taking advantage of a series of clock neuron subset-labeled drivers (Sekiguchi et al., 2020), we intersected CNMa-p65AD with 4 DN1 labelling drivers: GMR51H05-GAL4, GMR91F02-GAL4, Pdfr-KI-GAL4 and GMR79A11-GAL4 (Fig. 6E-6H). We found two arborization patterns: Type I with two neurons whose branches projecting to the anterior region, as observed in CNMa∩GMR51H05, CNMa∩Pdfr, and CNMa∩GMR79A11 (Fig 6E, 6G, 6H), and type II with four neurons branching on the posterior side with few projections to the anterior region, as observed in CNMa∩GMR91F02 (Fig. 6F). These two types of DN1ps’ subsets have been previously reported and profound discussed (Lamaze et al., 2018; Reinhard et al., 2022).

CNMa knockout in Type I or Type II neurons (GMR51H05-GAL4, GMR91F02-GAL4, and GMR79A11-GAL4) all reproduced the MAPI-increased phenotype of clk856 specific CNMa knockout (Fig. 6I). However, Type II neurons-specific CNMa knockout (CNMa ∩ GMR91F02) showed weaker advanced morning activity without advanced morning peak (Fig. 6N), while Type I neurons-specific CNMa knockout did (Fig. 6J), indicating a possibility that these two type I CNMa neurons are the main functional subset regulating the morning anticipation activity of fruit fly.

Pdf or Pdfr mutants exhibit weak or no morning anticipation, which is related to the phenotype of CNMa knockout flies. We also identified two Pdfr and CNMa double-positive DN1ps, which have a type I projection pattern (Fig. 6G). Reintroduction of Pdfr in Pdfr knockout background revealed that GMR51H05 and GMR79A11 Gal4 drivers, which covered the main functional CNMa-positive subset, could partially rescue the morning anticipation and power phenotype of Pdfr knockout flies to a considerably larger extent than the GMR91F02 driver (Fig. 6Q-6S, Fig. S13A, Table S7). Moreover, knocking out Pdfr by GMR51H05, GMR79A11 and CNMa GAL4, which cover type I CNMa neurons, decreased morning anticipation of flies (Fig. 6T, Fig. S13B). However, the decrease in morning anticipation observed in the Pdfr knockout by CNMa-GAL4 was not as pronounced as with the other two drivers. Because the presumptive main subset of functional CNMa is also PDFR-positive, there is a possibility that CNMa secretion is regulated by PDF/PDFR signal.

Role of Neuronal CNMaR in Morning Anticipation

There is only one CNMa receptor reported in the fly genome (Jung et al., 2014). We generated a CNMaRKO-p65AD line by CRISPR/Cas9 (Fig. 7A) and this knockout showed advanced morning activity (Fig. 7B, 7D) and increased MAPI (Fig. 7C, 7E) in both sexes. CNMaRKI-Gal4/UAS-mCD8::GFP and CNMaRKI-Gal4/UAS-stinger::Red showed expression of CNMaR across the whole brain (Fig. 7F, Fig S12B), especially in DN1p, DN3, the PI and the SOG. The dendrite arborization and synaptic projections of CNMaR neurons also covered broad regions (Fig. 7G, 7H), at the PI, the SOG, the posterior ventrolateral protocerebrum (PVLP) and the central complex (CC). Further conditional knockout of CNMaR in neurons by C-cCCTomics phenocopied CNMaRKO-p65AD phenotype (Fig. 7I-7L). These results indicate that CNMaR is similar to CNMa in regulating morning anticipation.

CNMaR Regulation of Morning Anticipation

(A) Schematic of CNMaRKO-p65AD generation. Most of the first exon in CNMaR was replaced by a T2A-p65AD-loxP-3P3-RFP-loxP cassette using CRISPR-Cas9 strategy and the T2A-p65AD was inserted in the reading frame of the remaining CNMaR codon. 3P3-RFP was removed latterly by Cre mediated recombination. (B-E) Activity plot (B, D) and statistical analysis (C, E) of male (B-C) or female (D-E) CNMaRKO-p65AD flies (red) and genotypical controls (blue and black). MAPI was significantly increased in both male and female CNMaRKO-p65AD flies. In this and other figures, “♀” denotes female flies. (F-H) Expression and projection patterns of CNMaR-KI-Gal4 in the brain. Scale bars, 50μm. (I-L) Activity plots (I, K) and statistical analyses (J, L) of CNMaR pan-neuronal knockout flies. Neuronal knockout of CNMaR increased MAPI (J, L).

Discussion

Conditional CCTomics Strategies and Toolkit

We have generated conditional gene manipulation systems based on Flp-out/GFPi or CRISPR/Cas9. cCCT based gene deletion after heat-shock or mifepristone (RU486) eliminated most GFP signals, and pan-neuronal constitutive expression of shRNAGFP or flippase disrupted seven out of eight tested genes completely while targeting of SIFaEGFP.FRT achieved 96±3% efficiency. Although the recombination of genetic elements is relatively cumbersome when using cCCTomics, it is worthwhile to apply this method to specific genes given its high level of efficiency. While two UAS-sgRNA libraries have been established, one primarily targeting kinases (Port et al., 2020) and the other targeting GPCRs (Schlichting et al., 2022), both libraries only cover a portion of CCT genes, and are thus insufficient for manipulating all CCT genes. The development of C-cCCTomics, however, makes CCT gene manipulation as simple as RNA interference. Furthermore, the use of modified Cas9.M6 or Cas9.M9 highly enhances the efficiency of gene disruption in the nervous system, allowing for efficient manipulation of all CCT genes in a cell-specific manner.

The toxicity of CRISPR/Cas9 depends on the Cas9 protein (Port et al., 2014). When expressed pan-neuronally in nSyb-GAL4 (R57C10-GAL4, attP40), Cas9.M9 slightly reduced viability while the expression of other Cas9 variants had no significant effect on viability (Fig. S14). Although Cas9.M9 showed leaky-expression-efficiency, this was not a problem with Cas9.M6, which successfully disrupted Dh31, Dh44, Pdf, and Pdfr (Fig. S9). A more restricted expression of Cas9.M9 with lower toxicity is necessary for better somatic gene manipulation in the future.

CCT of Clock Neurons

Intersecting Clk856-Gal4 or Clk856-p6AD with CCTomics, we identified 45 CCT genes in Clk856 labelled clock neurons. Clock neurons appear highly heterogeneous both in our intersection dissection and in a previous transcriptomic analysis (Abruzzi et al., 2017; Ma et al., 2021). Comparing these two CCT gene expression profiles in clock neurons, 41 out of 127 gene-subsets are identical. The accuracy of our genetic intersection is limited by two possibilities: 1) KI-LexA may not fully represent the expression pattern of the corresponding gene, and 2) the efficiency of STOP cassette removal in the Flp-out strategy is limited or the efficiency of LexA>LexAop::myrGFP. Moreover, the leakage of LexAop-GFP may result in unreliable labelling in split-lexA strategy. Both genetic drivers and transcriptomic analysis contribute to our knowledge of the expression profile of neurons. The physiological significance of each gene in particular neurons should be further investigated by genetic manipulation.

Regulation of Rhythmic Behavior by CCT genes

Multiple attractive genes have been identified in our functional screen of CCT genes in clock neurons: for example, knocking out of VGlut weakens morning anticipation (Table S8). In further screening of brain regions, we have narrowed down the morning anticipation regulation role of VGlut in R18H11-GAL4 labeled neurons (Table S9). VGlut in these neurons has also been reported to regulate sleep in Drosophila (Guo et al., 2016). Its downstream neurons may be the PI neurons or LNvs (Barber et al., 2021; Guo et al., 2016).

Moreover, the deficiency of the neuropeptide CNMa results in advanced morning activity. We have validated that two Pdfr and CNMa double-positive DN1p neurons may play a major role in regulating this process through intersectional manipulation of CNMa. Knockout and reintroduction of Pdfr in these neurons have verified that Pdfr partially functions in DN1p CNMa neurons, and PDF increases cAMP level in Pdfr positive neurons (Shafer et al., 2008), suggesting a possibility of the regulation of CNMa signaling by PDF signaling. Furthermore, given that the morning anticipation vanishing phenotype of Pdf or Pdfr mutant indicates a promoting role of PDF-PDFR signal, while the enhanced morning anticipation phenotype of CNMa mutant suggests an inhibiting role of CNMa signal, we consider the two signals to be antagonistic. However, knocking out CNMaR in Clk856 labelled clock neurons showed no significant phenotype (Table S6), whereas the mutant and pan-neuronal knockout flies had similar phenotypes to CNMa knockout flies, suggesting its role in the circadian output neurons. Previous studies have indicated that CNMa integrate thermosensory inputs to promote wakefulness, and CNMaR is thought to function in Dh44 positive PI neurons (Jin et al., 2021), a subset of circadian output neurons. To gain a deeper understanding of the downstream effects of DN1p CNMa positive neurons, further analysis focusing on specific brain regions is necessary.

We have also reproduced phenotypes of Pdf, Pdfr, and Dh31 mutant flies with C-cCCTomics as previous studies. Surprisingly, only five genes are functional among all sixty-seven CCT genes in this prior screen. This may be caused by limitations of the simple behavioral paradigm, single gene manipulation, and single GAL4 driver. For example, switching of light condition from L: D = 12h: 12h to L: D = 6h:18h, AstC/AstC-R2 would suppress flies’ evening activity intensity to adapt to the environmental change (Díaz et al., 2019), and only double knockout of AChRs and mGluRs in PI neurons can possibly result in alteration in behavioral rhythms (Barber et al., 2021). Further diversified functional analysis of CCT genes in clock neurons is required for clock circuit dissection.

Methods

Fly Lines and Rearing Conditions

Flies were reared on standard corn meal at 25 °C, 60% humidity, 12 h light: 12 h dark (LD) cycle. For flies used in behavior assays, they were backcrossed into our isogenized Canton S background for 5 to 7 generations. For heat induced assays, flies were reared at 20°C. All CCT attP KO lines and CCT KI driver lines were previous generated at our lab (Deng et al., 2019). Clk856-GAL4 and GMR57C10-GAL4 driver lines were gifts from Donggen Luo Lab (Peking University). 13XLexAop2 (FRT.stop) myr::GFP was gift from Rubin Lab.

C-cCCTomics sgRNAs design

All sgRNAs target at or before functional coding regions (e.g. GPCR transmembrane domain, synthetase substrate binding domain) of each CCT genes. For each gene, about 20 sgRNAs with specific score ≥ 12 were firstly designed at CRISPRgold website (Chu et al., 2016; Graf et al., 2019), then their specificity and efficacy were further valued in Optimal CRISPR target finder(Gratz et al., 2014), E-CRISPR(Heigwer et al., 2014) and CCTop (Stemmer et al., 2015) system. The first three highest efficacy sgRNAs with no predicted off-target effect were selected. All selected sgRNAs are listed in Table S3.

Molecular Biology

All cCCTomics knockin (KI) lines and C-cCCTomics transgenic flies were generated through phiC31 mediated attB/attP recombination, and the miniwhite gene was used as selection marker.

For cCCTomics KI lines, backbone pBSK-attB-FRT-HpaI-T2A-EGFP-FRT was modified from pBSK-attB-loxP-myc-T2A-Gal4-GMR-miniwhite (Deng et al., 2019). Myc-T2A-GAL4 cassette was removed by PCR amplification while first FRT cassette was introduced. Second FRT cassette was inserted by T4 ligation between SpeI and BamHI. T2A-EGFP was cloned from pEC14 and was inserted into the backbone between two FRT cassettes. All gene spans, except for stop codon, deleted in CCT attP KO lines were cloned into pBSK-attB-FRT-HpaI-T2A-EGFP-FRT at HpaI site.

For C-cCCTomics UAS-sgRNA lines, backbone pMsgNull was modified from pACU2 (Han et al., 2011). Synthetic partial fly tRNAGly sequence was inserted between EcoRI and KpnI. An irrelevant 1749 bp cassette amplified from pAAV-Efla-DIO-mScarlet (addgeen#130999) was inserted between EagI and KpnI. All sgRNA spacers were synthesized at primers, and “E+F” sgRNA scaffold and rice tRNAGly was amplified from a synthetic backbone PM04. Finally, gRNA-tRNAGly cassettes were cloned into pMsgNull between EagI and KpnI by Gibson Assembly.

All UAS-Cas9 variants generated in this research were cloned into vector pACU2 and all Cas9 sequence were amplified from hCas9 (addgene#41815). Human codon optimized Cas9 was cloned into pACU2 to generate UAS-Cas9.HC. UAS-Cas9.M0 was modified from UAS-Cas9.HC by introduce a 3x HA tag after NSL and replace the SARD linker with the 3×GGSGP linker (Zhao et al., 2016). UAS-Cas9.M6 and UAS-Cas9.M9 were designed as HMGN1-Cas9-UPD and HMGN1-Cas9-HB1 respectively. All these chromatin-modulating peptides were linked with Cas9 by 3x GGSGP linker.

CNMaRKO-p65AD is generated by replace the coding region of the first exon with T2A-p65AD by CRISPR/Cas9, the T2A-p65AD was linked in frame after first ten amino acids. Spacers of gRNAs used to break the targeted CNMaR region were 5’-GCAGATTTCAGTTCATCTTT-3’, 5’-GGCTTGGCAATGAC TATATA-3’.

Gene expression quantitation and high-throughput sequencing

Female flies were gathered six to eight days post-eclosion for gene expression quantification and high-throughput sequencing. Fly heads were isolated by chilling them on liquid nitrogen and subsequent shaking. mRNA extraction was performed using Trizol according to a previously established protocol (Green and Sambrook, 2020). Genomic DNA was removed, and cDNA was synthesized using a commercial kit (TIANGEN#DP419). For real-time quantitative PCR, at least one PCR primer was designed to overlap with the sgRNA target site.

Genomic DNA from fly heads was extracted using a standard alkali lysis protocol(Huang et al., 2009). Genomic regions approximately 130 to 230 bp in length, centered around the sgRNA target site, were amplified by PCR employing Q5 polymerase (NEB#M0494). Subsequently, libraries were prepared using the BTseq kit (Beijing Tsingke Biotech Co., Ltd.). These libraries were pooled and subjected to sequencing on the MiSeq platform (Illumina). Analysis of the libraries was conducted using Crispresso2(Clement et al., 2019).

Generation of KI and Transgenic lines

Generation of cCCTomics KI, CNMaKI-p65AD and CNMaRKO-p65AD lines are the same as generation of CCTomics KI driver lines as previously described (Deng et al., 2019). To generate C-cCCTomics UAS-sgRNA or UAS-Cas9 variants lines, attB vectors were injected and integrated into the attP40, attP2 or VK00005 through phiC31 mediated gene integration.

All flies generated in this research were selected by mini-white and confirmed by PCR.

Behavioral assays

Unmated male or female flies of 4 to 5 days were used in circadian rhythm assays. Before measurement, flies were entrained under 12 h light: 12 h dark cycle at 25°C for at least 3 days and then transferred to dark-dark condition for 7 days.

Virgin flies of 4 to 5 days were used in sleep assays. Flies were entrained to a 12 h light: 12 h dark cycle at 25°C for 2 days to eliminate the effect of CO2 anesthesia before sleep record. Sleep was defined as 5 min or longer immobility (Hendricks et al., 2000; Shaw et al., 2000) and analyzed by in-house scripts as previously described ((Dai et al., 2021; Dai et al., 2019; Deng et al., 2019).

Locomotion was obtained as previously described (Dai et al., 2021). Locomotion activity was measured and analyzed by Actogram J plugin (Dai et al., 2019). Morning anticipation index and evening anticipation index were defined as the ratio of last 3 hours activity before light-on or light-off accounts to last 6 hours activity before light-on or light-off, further subtracted by 0.5(Index=sum(3hrs)/sum(6hrs)-0.5) (Harrisingh et al., 2007; Im and Taghert, 2010; Seluzicki et al., 2014) and analyzed by an in-house python script.

Heat Shock and Drug Treatment

For hsFLP mediated conditional knockout, flies of 4 to 6 days were heat shock at 37°C during ZT10 to ZT12 for 4 days. And they were reared at 20°C for another 4-days and then dissected.

For mifepristone (RU486) induced conditional knockout, flies of 4 to 6 days were treated with 500 μM RU486 mixed in corn food and then dissected 4 days later.

Immunohistochemistry and Confocal Imaging

For all imaging without staining, adult flies were anesthetized on ice and dissected in cold phosphate-buffer saline (PBS). Brains or ventral nerve codes (VNC) were fixed in 2% paraformaldehyde (weight/volume) for 30 min, washed with washing buffer (PBS with 1% Triton X-100, v/v, 3% NaCl, g/mL) for 7 min three times and mounted in Focusclear (Cell Explorer Labs, FC-101).

For imaging with staining, brains and VNCs were fixed for 30 min, washed for 15 min three times. Then they were blocked in PBSTS, incubated with primary antibodies, washed with washing buffer, incubated with second antibodies and mounted as described previously(Dai et al., 2021; Dai et al., 2019).

All brains or VNCs were imaged on Zeiss LSM710 or Zeiss LSM880 confocal microscope and processed by Imaris.

The following primary antibodies were used: mouse anti-PDF (1:200, DSHB), rabbit anti-TH (1:1000, Novus Biologicals), rabbit anti-LK (1:1000, RaoLab, this paper). Rabbit anti-DSK (1:1000) was a gift from Dr. C. Zhou Lab (Institute of Zoology, Chinese Academy of Science) (Wu et al., 2020). The following secondary antibodies were used: AlexaFluor goat anti-mouse 488 (1:1000, Invitrogen), AlexaFlour goat anti-rabbit 488/633 (1:1000, Invitrogen).

For Fig 2, number of TH positive neurons were counted with Imaris Spots plugin.

Quantification and Statistics

MAI, MAPI, EAI, EAPI, power and period were calculated by python or R scripts. ZT0 was set as the time point when light was on and ZT12 was set as the time point for light-off. Activity bins started at ZT0 and each was calculated as a sum of the total activity within 30min. Flies were regarded as death and removed if their activity value within last 2 bins was 0. A representative 24hr activity pattern was the average between corresponding activity bins from 2 consecutive days. To minimize effects from singular values, each flies’ activity was normalized using the following formula: bi means the activity value for a certain bin. min(b0,…,b48) means the minimal bin value within 24hr and max(b0,…,b48) means the maximal bin value within 24hr. Nor_bi means the final normalized bin value for a certain bin from a given fly. Normalized activity was used for following analysis.

Morning activity arise (M_arise) was defined as the radian between the activity curve (ZT21-ZT22) and the time coordinate. Morning activity plateau (M_plateau) was defined as the radian between the activity curve (ZT22.5-ZT24) and the time coordinate. Evening activity arise (E_arise) was defined as the radian between the activty curve (ZT8-ZT11.5) and the time coordinate. Evening activity plateau (E_plateau) was defined as the radian between the activity curve (ZT10.5-ZT12) and the time coordinate. Morning anticipation pattern index (MAPI) was calculated by subtracting M_arise from M_plateau and Evening anticipation pattern index (EAPI) was calculated by subtracting E_arise from E_plateau.

The original activity data from 7 consecutive days in dark-dark condition was used for power and period calculation as described (Geissmann et al., 2019). Each flies’ periodogram was calculated based on Chi-Square algorithm (Sokolove and Bushell, 1978) and flies with a null power value were regarded as arrhythmic.

All statistical analyses were carried out with Prism 8 (Graphpad software). The Kruskal-Wallis ANOVA test followed by Dunn’s posttest was used to compare multiple columns.